Therapeutic benefits of maintaining CDK4/6 inhibitors and incorporating CDK2 inhibitors beyond progression in breast cancer

Version 1

Abstract

Abstract

The combination of CDK4/6 inhibitors (CDK4/6i) and endocrine therapy has revolutionized treatment for hormone receptor-positive (HR+) metastatic breast cancer. However, the emergence of resistance in most patients often leads to treatment discontinuation with no consensus on effective second-line therapies. The therapeutic benefits of maintaining CDK4/6i or incorporating CDK2 inhibitors (CDK2i) after disease progression remain unclear. Here, we demonstrate that sustained CDK4/6i therapy, either alone or combined with CDK2i, significantly suppresses the growth of drug-resistant HR+ breast cancer. Continued CDK4/6i treatment induces a non-canonical pathway for retinoblastoma protein (Rb) inactivation via post-translational degradation, resulting in diminished E2F activity and delayed G1 progression. Importantly, our data highlight that CDK2i should be combined with CDK4/6i to effectively suppress CDK2 activity and overcome resistance. We also identify cyclin E overexpression as a key driver of resistance to CDK4/6 and CDK2 inhibition. These findings provide crucial insights into overcoming resistance in HR+ breast cancer, supporting the continued use of CDK4/6i and the strategic incorporation of CDK2i to improve therapeutic outcomes.

Introduction

Metastatic breast cancer remains a leading cause of cancer-related mortality in women globally (1,2). A key dysregulation in breast cancer involves the overactivation of cyclin-dependent kinases 4 and 6 (CDK4/6) (3–5). Active CDK4/6 phosphorylates the retinoblastoma protein (Rb), a crucial regulator that prevents cell-cycle initiation by sequestering E2F transcription factors (6). Rb phosphorylation results in the release of E2F, thereby promoting CDK2 activation and cell proliferation (7,8). Understanding this mechanism has driven significant advancements in therapeutic strategies, particularly for hormone receptor-positive (HR+)/human epidermal growth factor receptor 2-negative (HER2-) breast cancer, which constitutes approximately 70% of breast cancer cases (9). The current standard first-line treatment for HR+/HER2- metastatic breast cancer is a combination of CDK4/6 inhibitors (CDK4/6i) and endocrine therapy (ET) (4,5,10). Although this approach has markedly improved patient outcomes, resistance remains a significant challenge, with about 30% of patients developing resistance within two years and most eventually facing resistance (11,12). Upon disease progression, CDK4/6i therapy is often discontinued, leading to aggressive tumor growth with effective second-line therapies lacking.

While continuing ET after disease progression has demonstrated clinical benefits (13–17), the efficacy of maintaining CDK4/6i treatment post-progression is still elusive. Ongoing clinical trials investigate the potential advantages of continuing CDK4/6i therapy using FDA-approved inhibitors such as palbociclib, abemaciclib, and ribociclib in patients experiencing disease progression (18–22). The MAINTAIN phase 2 trial suggests that continuing ribociclib with alternative ET post-progression on a palbociclib and ET regimen can significantly extend progression-free survival (PFS) compared to switching to ET alone (22). Similarly, interim results from the postMONARCH phase 3 study indicate that adding abemaciclib to ET improved PFS compared to ET alone in patients who have progressed on ET and a prior CDK4/6i (23). Conversely, the PACE and PALMIRA clinical trials did not observe a significant PFS benefit when maintaining palbociclib combined with new ET after disease progression (19,24).

Like CDK4/6, CDK2 also phosphorylates Rb before the G1/S transition (7,25). The relevance of CDK2 activation in CDK4/6i-resistant tumors highlights the potential of targeting CDK2 to overcome CDK4/6i resistance (26–30). Several clinical trials are exploring CDK2 inhibitors (CDK2i) in patients who have progressed on CDK4/6i-based treatments as monotherapies or in combination with CDK4/6i. Determining the optimal strategy for incorporating CDK2i in treating HR+/HER2- breast cancer that has developed drug resistance remains a critical question.

This study emphasizes that continued CDK4/6i treatment in drug-resistant cells leads to incomplete and ineffective retinoblastoma (Rb) inactivation, significantly slowing G1-phase progression. Additionally, it highlights the therapeutic potential of combining CDK2i with CDK4/6i as an effective second-line treatment strategy. Notably, the research underscores the critical role of cyclin E overexpression in driving resistance to the CDK2i and CDK4/6i combination, providing insights into overcoming such resistance in future cancer treatments.

Results

Maintaining CDK4/6i treatment attenuates the growth of drug-resistant cells by extending G1-phase progression

To investigate the effect of maintaining CDK4/6i treatment on drug-resistant cells, we used three HR+/HER2- breast cancer cell lines (MCF-7, T47D, and CAMA-1) and a triple-negative breast cancer cell line (MDA-MB-231). These cells were chronically exposed to palbociclib for over a month to induce drug resistance. We confirmed increased half-maximal inhibitory concentrations (IC50) for palbociclib in drug-resistant cells compared to drug-naïve cells (Figure S1A). We either maintained or discontinued CDK4/6i treatment in these drug-resistant cells and evaluated the cumulative proliferation rate over 12 days, along with drug-naïve cells. We found a significantly slower growth rate in drug-resistant cells maintained on CDK4/6i treatment compared to other conditions (Figure 1A). Conversely, drug-resistant cells with discontinued CDK4/6i treatment exhibited similar growth rates to drug-naïve cells. These results suggest that despite the development of drug resistance, maintaining CDK4/6i treatment significantly impedes the growth rate of drug-resistant cells.

Continuous CDK4/6i treatment attenuates drug-resistant cell growth.

(A) Growth curves of drug-naïve and drug-resistant cells. Palbociclib (1 µM) was either withdrawn or maintained in drug-resistant cells. Data are shown as mean ± SD (n = 3 biological replicates). Statistical significance was determined using one-way ANOVA with Tukey’s post-hoc analysis (**** P < 0.0001). (B) Schematic illustration of live-cell sensors for CDK4/6 and CDK2 activities (top) and cell-cycle phase (bottom). (C-D) Representative single-cell traces showing CDK4/6 and CDK2 activities and Cdt1-degron intensity in drug-naïve (C) and drug-resistant (D) MCF-7 cells. Palbociclib (1 µM) was maintained in drug-resistant cells. Vertical gray bars indicate mitotic events, while circles mark G1/S (red) and S/G2 (blue) transitions. (E-F) Heatmaps showing single-cell traces for CDK4/6 (left) and CDK2 (middle) activities, and Cdt1-degron intensity (right) in drug-resistant cells without (E) or with (F) continuous palbociclib (1 µM) treatment. Proliferating cells were identified based on CDK2 activity (>1 for more than 2 hr between 30 and 48 hr). (G-H) Single-cell traces of CDK2 activity (left) and Cdt1-degron intensity (right) aligned by the end of mitosis (anaphase) in drug-resistant cells without (G) or with (H) continuous palbociclib (1 µM) treatment. Based on CDK2 activity (black line), cells were classified as either proliferation (red) or quiescent (blue). Mitosis timing is marked in gray.

We next monitored cell-cycle progression in individual cells to understand how continued CDK4/6i treatment suppresses the growth rate in drug-resistant cells. We employed MCF-7 and MDA-MB-231 cells expressing live-cell sensors for CDK4/6 (31) and CDK2 (32) activities, cell-cycle phases (Cdt1 degron) (33), and a nucleus marker (histone 2B) for individual cell tracking. The CDK sensors dynamically shuttle between the nucleus and cytoplasm based on their phosphorylation by the relevant kinase (Figure 1B top). The Cdt1 degron is degraded during the S phase, enabling visualization of the G1/S and S/G2 transitions (Figure 1B bottom). After mitosis, proliferating drug-naïve cells continuously activated CDK4/6, followed by gradual increase in CDK2 activity to enter the S phase (Figure 1C). In contrast, CDK4/6i-resistant cells activated CDK2 with low CDK4/6 activity to enter the S phase (Figure 1D). When evaluating thousands of cells, approximately 98% of drug-naïve cells continuously activated both CDK4/6 and CDK2 to enter the cell cycle (Figure S1B and S1C). While drug withdrawal led to CDK4/6 reactivation, over 90% of drug-resistant cells continued to proliferate regardless of the presence of CDK4/6i (Figure 1E, 1F, S1D, and S1E). To monitor cell-cycle dynamics, we aligned cells by the end of mitosis (cytokinesis) and classified them based on CDK2 activity into proliferating and quiescent cells. Without CDK4/6i treatment, drug-resistant cells exhibited similar kinetics of CDK4/6 and CDK2 activities, as well as S-phase entry, compared to drug-naïve cells (Figure 1G and S2A–S2D). However, with continued CDK4/6i treatment, drug-resistant cells showed low CDK4/6 activity, slower activation kinetics of CDK2, and increased heterogeneity in the timing of S-phase entry, as indicated by Cdt1 degradation, compared to other conditions (Figure 1H, S2E, and S2F). These results indicate that continuous treatment with CDK4/6i leads to marked deceleration and irregularity in cell-cycle progression of drug-resistant cells, thus suppressing their growth rates.

CDK6 overexpression has emerged as a key driver of CDK4/6i resistance (34–37). In line with these findings, we observed elevated CDK6 levels in CDK4/6i-resistant cells (Figure S3A). To investigate the role of CDK6 in modulating CDK4/6 activity under resistance, we established CDK6 knockout (KO) MCF-7 cells using CRISPR-Cas9 and induced resistance to CDK4/6i (Figure S3B). Despite CDK6 depletion, CDK4/6 activity remained comparable between wild-type and CDK6-KO cells under continuous CDK4/6i treatment (Figure S3C–S3F). These results suggest a potential non-canonical role for CDK6 in mediating resistance to CDK4/6i therapy.

To assess the impact of continuous CDK4/6i treatment on cell-cycle progression, we measured the intermitotic time and the duration of each cell-cycle phase in MCF-7 and CAMA-1 cells (Figure 2A). The intermitotic time was comparable between drug-naïve and drug-resistant cells without ongoing CDK4/6i treatment (Figure 2B). However, in drug-resistant cells that were continuously treated with CDK4/6i, the intermitotic time was extended by 30–50% compared to other conditions. Moreover, continuous CDK4/6i treatment prolonged the G1 phase by approximately 200–300% without affecting the duration of the S and G2/M phases (Figure 2C–2E). Our data indicate that maintaining CDK4/6i treatment in drug-resistant cells significantly extends the G1 phase, thereby decelerating the overall growth rate.

CDK4/6i maintenance extends G1-phase progression.

(A) Representative single-cell traces showing CDK4/6 and CDK2 activities (top) and Cdt1-degron intensity (bottom) in drug-naïve MCF-7 cells. (B-E) Violin plots showing intermitotic time (n > 600 cells/condition) (B), G1-phase duration (n > 2,000 cells/condition) (C), S-phase duration (n > 400 cells/condition) (D), and G2/M-phase duration (n > 250 cells/condition) (E) in MCF-7 (left) and CAMA-1 (right) cells. Solid and dashed yellow lines indicate mean and median, respectively.

Maintaining CDK4/6i treatment triggers an ineffective Rb inactivation pathway

We sought to elucidate the molecular mechanisms underlying the extended G1 phase and slow CDK2 activation kinetics in drug-resistant cells under continuous CDK4/6i treatment. Our recent studies demonstrated that CDK4/6 inhibition initially halts cell proliferation through Rb activation but ultimately reduces Rb protein due to decreased stability (38,39). We confirmed a decrease in both total and phosphorylated Rb levels in drug-resistant cells continuously treated with CDK4/6i compared to drug-naïve cells (Figure 3A). Discontinuation of CDK4/6i treatment restored both total and phosphorylated Rb levels. Importantly, the reduction of Rb protein in drug-resistant cells was incomplete, as evidenced by comparisons to Rb-KO cells (Figure S4A). Furthermore, Rb KO in drug-resistant cells reversed the extended G1 duration and the slowed activation kinetics of CDK2 without affecting CDK4/6 activity, even in the presence of continuous CDK4/6i treatment (Figure S4B and S4C). This led us to hypothesize that maintaining CDK4/6i treatment in drug-resistant cells induces a non-canonical pathway for Rb inactivation via its degradation. This passive Rb inactivation may result in ineffective E2F and CDK2 activation kinetics, delaying G1-phase progression.

CDK4/6i maintenance triggers an ineffective Rb inactivation pathway.

(A) Immunoblot showing phosphorylated Rb (S807/811) (p-Rb), total Rb (t-Rb), and β-actin protein levels in MCF-7 cells. Drug-resistant cells were harvested two weeks after drug withdrawal. (B) Singe-cell traces showing CDK2 activity aligned by mitosis in proliferating (red) and quiescent (blue) MCF-7 cells. Circles indicate the time of fixation and staining (n = 200 cells). (C) Scatter plot of EdU intensity classified by time since mitosis in drug-naïve cells. The red dotted line marks the threshold for S phase (n = 2,000 cells). (D-E) Percentage of S-phase cells as a function of time since mitosis in proliferating (D) and quiescent (E) cells. Data are shown as mean ± SD (n = 2 biological replicates). (F) Representative immunostaining of Hoechst, p-Rb, t-Rb, and E2F1 mRNA FISH (top). Processed images that show nucleus segmentation, p-Rb classification, and mRNA puncta detection (bottom). (G) Histogram showing the percentage of cells with p-Rb normalized by t-Rb. (H-I) Percentage of cells with p-Rb (H) and E2F1 mRNA levels (I) as a function of time since mitosis. Data are shown as mean ± SD (n = 3 biological replicates) (H) or mean ± 95% confidence intervals (n > 2,500 cells/condition) (I). (J-K) E2F1 mRNA (J) and EdU (K) levels as a function of CDK2 activity. Data are shown as mean ± 95% confidence intervals (n > 2,500 cells/condition).

To test our hypothesis, we monitored CDK2 activity to isolate proliferating cells and aligned them with corresponding fixed-cell data by the end of mitosis (Figure 3B and S5A). Using 5-ethynyl-2’-deoxyuridine (EdU) incorporation, we measured the kinetics of entry into the S phase. Almost all drug-naïve and -resistant cells without CDK4/6i treatment transitioned sharply into the S-phase about 6 hr after mitosis and completed DNA replication within 12 hr (Figure 3C, 3D, and S5B). However, drug-resistant cells under continued CDK4/6i treatment exhibited a dramatically slowed G1/S transition and increased variability in entering the S-phase compared to other conditions (Figure 3D and S5C). We also confirmed no S-phase entry in quiescent cells (Figure 3E). Using immunostaining and mRNA FISH, we evaluated the kinetics of Rb phosphorylation and E2F activity (Figure 3F). Cells were classified based on Rb phosphorylation at Serine 807/811, which is a previously characterized marker for hyperphosphorylation (40) (Figure 3G). The expression levels of E2F1 mRNA are regulated via autoregulatory control (41), thereby serving as a proxy of E2F transcriptional activity (7,40). In the absence of CDK4/6i, both drug-naïve and -resistant cells fully induced Rb phosphorylation before S-phase entry (Figure 3H). Conversely, in the presence of CDK4/6i, drug-resistant cells initiated CDK2 activation and G1 progression without Rb phosphorylation, showing low and slow E2F mRNA accumulation (Figure 3H and I). Moreover, when comparing CDK2 activity with levels of E2F1 mRNA or EdU incorporation, drug-resistant cells under continuous CDK4/6i treatment required higher CDK2 activity to achieve the same levels of E2F1 mRNA and to trigger the G1/S transition compared to other conditions (Figure 3J, 3K, S5D, and S5E). The observed reduction in E2F1 mRNA during the S phase could be attributed to the suppression of atypical E2Fs, E2F7 and E2F8 (42). These findings suggest that maintaining CDK4/6i treatment in drug-resistant cells induces inefficient Rb inactivation, characterized by reduced and delayed E2F activation kinetics. Such alterations substantially increase the duration of the G1 phase and the variability in the G1/S transition.

Maintenance of CDK4/6i suppresses the growth of drug-resistant tumors

We investigated the therapeutic benefits of maintaining CDK4/6i treatment after the onset of drug resistance in vivo. To this end, we established an MCF-7 xenograft model by orthotopically inoculating drug-naïve MCF-7 cells into the mammary fat pad of immunodeficient mice. When the tumor reached a volume of 100 mm3, we treated mice with palbociclib and monitored tumor growth (Figure 4A). After an initial phase of tumor stasis and regression lasting for varying durations, we observed the emergence of drug resistance, evidenced by tumor regrowth (Figure 4B). Once tumors regrew to volumes between 145 and 155 mm3, the mice were randomly assigned to secondary treatment groups: discontinuation of drug treatment, continuation with palbociclib, or switching to ribociclib or abemaciclib. Consistent with our in vitro results, maintaining CDK4/6i significantly attenuated overall tumor growth compared to discontinuing drug treatment (Figure 4C–E). Analysis of Rb expression in tumor tissues post-secondary treatment revealed the restoration of total Rb expression exclusively in the drug-discontinued condition (Figure 4F). These data underscore the therapeutic advantages of maintaining CDK4/6i to attenuate the growth of drug-resistant tumors.

CDK4/6i maintenance suppresses the growth of drug-resistant tumors.

(A) Schematic representation of experimental design. Once tumors reached a volume of 100 mm3, mice were treated with palbociclib. Following the development of resistance, mice were randomly assigned to one of four treatment groups: treatment discontinuation, palbociclib maintenance, switch to ribociclib, or switch to abemaciclib for 32 days. (B) Tumor growth curves showing the establishment of resistance to palbociclib. Horizontal red dotted lines (145−155 mm3) indicate the point at which mice were assigned to second-line treatment. (C-D) Averaged (C) and individual (D) tumor growth traces following second-line treatments. Data are shown as mean ± SEM. Statistical significance was determined using one-way ANOVA with Tukey’s post-hoc analysis (* P < 0.05, **** P < 0.0001). (E-F) Box plots showing tumor mass (E) and the percentage of Rb-positive cells (F) after second-line treatment. Statistical significance was determined using one-way ANOVA with Tukey’s post-hoc analysis (** P < 0.05, *** P < 0.01, **** P < 0.0001).

Addition of ET augments the efficacy of CDK4/6i maintenance

Our recent studies have highlighted the pivotal role of c-Myc in amplifying E2F activity and fostering CDK4/6i resistance following alternative Rb inactivation (38,39). Using a doxycycline-inducible system, we found that induction of c-Myc significantly facilitated the growth of drug-resistant cells under ongoing CDK4/6i treatment (Figure S6A). Furthermore, despite similar low CDK4/6 activity, c-Myc induction facilitated cell-cycle progression in CDK4/6i-resistant cells (Figure S6B–D). These data demonstrated that c-Myc promotes CDK4/6i resistance by facilitating cell-cycle progression in drug-resistant cells.

Given the estrogen responsiveness of c-Myc (43), we hypothesized that the continued addition of ET could further suppress the growth of CDK4/6i-resistant cells by downregulating c-Myc expression. We employed the estrogen receptor antagonist fulvestrant to evaluate the therapeutic benefit of maintaining the combination of CDK4/6i and ET. While MCF-7 cells exhibited primary resistance to fulvestrant, CAMA-1 and T47D cells displayed sensitivity to this drug (Figure 5A). We used these MCF-7 and CAMA-1 cells to examine whether primary ET resistance contributes to the benefit of maintaining CDK4/6i and ET in drug-resistant cells. These cells were chronically exposed to palbociclib and fulvestrant for over two months to induce drug resistance. Subsequently, these resistant cell lines were subjected to continuous treatment with either palbociclib or fulvestrant alone or in combination, and their cumulative proliferation rate was monitored. We observed a significantly slower growth rate in drug-resistant cells continuously exposed to the combination therapy compared to other treatment conditions, not only in CAMA-1 but also in MCF-7 cells (Figure 5B, S7A, and S7B). Continued treatment with fulvestrant, either alone or combined with palbociclib, reduced c-Myc levels compared to palbociclib maintenance alone (Figure 5C). Our data indicate distinct resistance mechanisms between CDK4/6i and ET. Therefore, regardless of primary resistance to ET, the continued addition of ET to CDK4/6i treatment is beneficial in suppressing the growth of drug-resistant cells by inhibiting c-Myc.

Maintaining CDK4/6i and ET synergistically suppresses the growth of drug-resistant cells.

(A) IC50 values of fulvestrant. Data are shown as mean ± SD (n = 3 biological replicates). Statistical significance was determined using one-way ANOVA with Tukey’s post-hoc analysis (*** P < 0.001). (B) Growth curves of MCF-7 (left) and CAMA-1 (right) cells resistant to palbociclib and fulvestrant. Cells were maintained with fulvestrant (500 nM) or palbociclib (1 µM) alone or their combination. Data are shown as mean ± SD (n = 3 biological replicates). Statistical significance was determined using one-way ANOVA with Tukey’s post-hoc analysis (** P < 0.01, *** P < 0.001, **** P < 0.0001). (C) Violin plots representing c-Myc levels in cells resistant to palbociclib and fulvestrant treated with the indicated drugs for one week before fixation. Solid and dashed yellow lines represent mean and median, respectively (n > 2,000 cells/condition). (D-E) Heatmaps showing single-cell traces for CDK4/6 (left) and CDK2 (middle) activities and Cdt1-degron intensity (right) in drug-resistant cells maintained with palbociclib (1 µM) and fulvestrant (500 nM) (D) or fulvestrant alone (E). Proliferating cells were identified based on CDK2 activity (>1 for more than 2 hr between 30 and 48 hr). (F) Averaged traces of CDK4/6 (left) and CDK2 (right) activities aligned by mitosis in drug-resistant cells treated with the indicated drugs. Data are shown as mean ± 95% confidence intervals (n > 1,000 cells/condition). (G) Single-cell traces of Cdt1-degron intensity aligned by mitosis in drug-resistant cells treated with palbociclib (1 µM) (left) or palbociclib + fulvestrant (500 nM) (right). (H-I) Violin plots showing intermitotic time (n > 74 cells/condition) (H), and G1-phase duration (n > 630 cells/condition) (I) in MCF-7 (left) and CAMA-1 (right) cells. Solid and dashed yellow lines indicate mean and median, respectively.

To elucidate the mechanisms underlying the attenuation of growth resulting from continued treatment with the drug combination, we assessed CDK4/6 and CDK2 activities along with cell-cycle progression (Cdt1 degron) in drug-resistant MCF-7 cells. Cells were continuously treated with palbociclib or fulvestrant alone or in combination. By classifying cells into proliferation and quiescence based on CDK2 activation, we found over 80% of cells were proliferating in all conditions (Figure 5D, 5E, and S7C). Furthermore, we observed reactivation of CDK4/6 in drug-resistant cells continuously treated with fulvestrant alone. However, when we monitored cell-cycle dynamics by aligning cells to mitosis, the maintenance of the drug combination caused slower CDK2 activation kinetics and greater heterogeneity in S-phase entry, as indicated by Cdt1 degradation, compared to single-drug treatment conditions (Figure 5F, 5G, S7D, and S7E). By analyzing each cell-cycle duration, we observed a significant increase in intermitotic time and G1-phase duration, while S- or G2-phase durations remained relatively unchanged (Figure 5H, 5I, S7F, and S7G). This emphasizes the role of the drug combination in altering cell-cycle kinetics in the G1 phase. Our findings showed the therapeutic benefits of maintaining a combination of CDK4/6i and ET in attenuating the growth of drug-resistant cells.

The benefit of using CDK2i in combination with CDK4/6i as a second-line therapy

We next evaluated the potential of integrating CDK2i as a second-line treatment by employing the CDK2i INX-315. We examined the growth of palbociclib/fulvestrant-resistant MCF-7 and CAMA-1 cells under various treatment regimens: treatment discontinuation, fulvestrant alone, fulvestrant + palbociclib, fulvestrant + INX-315, and fulvestrant + palbociclib + INX-315. Continued treatment with the combination of palbociclib and fulvestrant significantly reduced the growth rate of drug-resistant cells compared to treatment discontinuation, fulvestrant alone, and fulvestrant + INX-315 (Figure 6A–6C). Notably, the triple combination most effectively suppressed the growth of these drug-resistant cells. We next investigated cell-cycle dynamics by assessing CDK activities and Cdt1-degron levels a week after drug treatment. Consistent with earlier results, CDK4/6 reactivation was observed following palbociclib withdrawal (Figure S8A–S8D). Furthermore, we found CDK2 reactivation even in the presence of INX-315 (Figure 6D and S8D). The triple-drug combination caused the most extended intermitotic time and G1-phase duration without affecting the S and G2 phases (Figure 6E and S8E). Analysis of cells synchronized by mitosis revealed that the triple-drug combination most prominently delayed CDK2 activation kinetics (Figure 6F). To further understand the effects of second-line therapies on the transcriptome landscape of CDK4/6i/ET-resistant cells, we performed RNA sequencing 20 days after drug treatment. In line with our previous observations, continued palbociclib and fulvestrant more effectively suppressed E2F-target gene expression compared to the INX-315 and fulvestrant combination, with the triple-drug regimen showing the most robust inhibition of both E2F and Myc target gene expression (Figure 6G and S8F). These data suggest that despite CDK2 reactivation, the addition of CDK2i to CDK4/6i and ET maintenance therapy further attenuates E2F activity and CDK2 activation kinetics, thereby inhibiting the growth rate of drug-resistant cells.

Therapeutic benefit of combining CDK2i with CDK4/6i and ET.

(A-B) Growth curves of MCF-7 (left) and CAMA-1 (right) cells resistant to palbociclib and fulvestrant under various treatment conditions: drug discontinuation, fulvestrant (500 nM) alone, or in combination with palbociclib (1 µM) and/or INX-315 (100 nM). Data are shown as mean ± SD (n = 3 biological replicates). P values were calculated with an unpaired t-test (A) and using one-way ANOVA with Tukey’s post-hoc analysis (B) (*P < 0.05, ** P < 0.01, *** P < 0.001, **** P < 0.0001). (C) Cell numbers 20 days after drug treatment. Data are shown as mean ± SD (n = 3 biological replicates). P values were calculated using an unpaired t-test (* P < 0.05, ** P < 0.01, *** P < 0.001, **** P < 0.0001). (D) Heatmaps of single-cell traces for CDK4/6 (left) and CDK2 (middle) activities, and Cdt1-degron intensity (right) in palbociclib/fulvestrant resistant MCF-7 cells treated with the triple combination of palbociclib (1 µM), fulvestrant (500 nM), and INX-315 (100 nM) for one week before imaging. Proliferating cells were identified based on CDK2 activity (>1 for more than 2 hr between 30 and 48 hr). (E) Violin plots showing intermitotic time (n > 200 cells/condition) (left) and G1-phase duration (n > 900 cells/condition) (right). Solid and dashed yellow lines indicate mean and median, respectively. (F) Averaged traces of CDK4/6 (left) and CDK2 (right) activities aligned by mitosis in drug-resistant cells treated with the indicated drugs. Data are shown as mean ± 95% confidence intervals (n > 1,000 cells/condition). (G) Heatmaps comparing gene expression profiles for hallmark E2F (left) and MYC (right) targets in drug-resistant MCF-7 cells treated with the indicated drugs for 20 days. Samples were collected as biological duplicates.

Role of cyclin E/A in CDK2 reactivation under combined CDK4/6 and CDK2 inhibition

A recent study identifies the cyclin A-CDK complex as a key driver of rapid adaptation to CDK2 inhibition (44). To investigate the mechanism of CDK2 reactivation during CDK2i treatment, we explored the role of the CDK2 activators, cyclin E and A. Using a doxycycline-inducible system, we induced the overexpression of cyclin E1 or A2 in drug-naïve MCF-7 cells (Figure S9A). While treatment with INX-315 (100 nM–1 µM) initially inhibited CDK2 activity, CDK2 reactivation occurred within 1–2 hr after drug treatment, regardless of cyclin E1 or A2 overexpression (Figure 7A and S9B–S9D). The combination of CDK2i and CDK4/6i blocked this rapid CDK2 reactivation, but the overexpression of cyclin E1 or A2 hindered complete suppression (Figure 7B and S9E–S9G). These findings suggest that CDK4/6 inhibition is necessary for effectively suppressing CDK2 activity by CDK2i, and that cyclin E1 and A2 overexpression may contribute to CDK2 reactivation.

Overexpression of cyclin E and A facilitates resistance to the combination of CDK2i and CDK4/6i.

(A-B) Averaged traces of CDK4/6 (left) and CDK2 (right) activities in MCF-7 cells with or without cyclin E1 or A2 overexpression. Cells were treated with doxycycline (500 nM) 6 hr before the addition of INX-315 (1 µM) alone (A) or in combination with palbociclib (1 µM) (B). Data are shown as mean ± 95% confidence intervals (n > 900 cells/condition). (C-D) Heatmaps of single-cell traces for CDK4/6 (left) and CDK2 (middle) activities, and Geminin-degron intensity (right) in drug-naïve MCF-7 cells without (C) or with (D) cyclin E1 overexpression. Cells were treated with palbociclib (1 µM), INX-315 (100 nM), and doxycycline (500 nM). Persister cells were identified based on CDK2 activity (>1 for more than 2 hr between 30 and 48 hr). (E) Averaged traces of CDK4/6 (left) and CDK2 (middle) activities and Geminin degron intensity (right) in persister cells tolerant to combination palbociclib (1 µM), and INX-315 (100 nM) without or with cyclin E1 or A2 overexpression. Data are shown as mean ± 95% confidence interval. n > 2500 cells/condition. (F) Growth curves of palbociclib-resistant MCF-7 cells without or with cyclin E1 or A2 overexpression. Cells were treated with palbociclib (1 µM), INX-315 (100 nM), and doxycycline (500 nM). Data are shown as mean ± SD (n = 3 biological replicates). Statistical significance was determined using one-way ANOVA with Tukey’s post-hoc analysis (** P < 0.01, **** P < 0.0001). (G) Summary schematic illustrating the mechanisms underlying the benefits of continued CDK4/6i and ET therapies and the introduction of CDK2i in drug-resistant cells.

We next examined the impact of cyclin E and A overexpression in the development of persister cells, which are implicated in residual tumor growth and eventual resistance (38,45). We used drug-naïve MCF-7 cells expressing live-cell sensors for CDK2, CDK4/6, and anaphase-promoting complex/cyclosome (APC/C) (Geminin degron) activities (46). APC/C is a multi-subunit E3 ubiquitin ligase typically inactivated near the G1/S transition, leading to Geminin-degron accumulation. Treatment with CDK4/6i and CDK2i induced near-complete cell-cycle arrest within 24 hr (Figure 7C). However, approximately 10% of cells developed a persister phenotype through CDK2 reactivation. Overexpression of cyclin E1 increased the percentage of persister cells, while cyclin A2 had no obvious impact (Figure 7D and S10). Further kinetic analysis revealed that cyclin E1 overexpression accelerated CDK2 reactivation and APC/C inactivation in persister cells (Figure 7E). These findings indicate that cyclin E1, but not cyclin A2, facilitates CDK2 reactivation and persister development, promoting resistance to the CDK4/6i and CDK2i combination.

Since cyclin A is a substrate of APC/C (47), the lack of effect from cyclin A2 induction could be due to its degradation by APC/C. Thus, this degradation may reduce cyclin A to insufficient levels, preventing its impact on persister cell development. To further evaluate the impact of cyclin E and A overexpression on resistance development, we exposed palbociclib-resistant MCF-7 cells to the combination of palbociclib and INX-315. Both cyclin E1 and A2 overexpression significantly accelerated the development of resistance to the drug combination, with cyclin E1 having a more significant effect than cyclin A2 (Figure 7F). These results indicate that although CDK2 activity is targeted by CDK2i, the overexpression of cyclin E and A promotes resistance to the combination of CDK4/6i and CDK2i. Due to cyclin A degradation by APC/C in quiescent cells, cyclin E may play a more critical role in driving resistance to this drug combination.

Discussion

The combination therapy of CDK4/6i and ET has revolutionized the treatment landscape for HR+/HER2- breast cancer (48–55). Despite the initial efficacy of these treatments, the inevitable emergence of drug resistance remains a significant challenge, mainly as there is no established second-line therapy. Our findings demonstrate that sustained CDK4/6i treatment in drug-resistant cells results in ineffective Rb inactivation, significantly delaying E2F activation and G1-phase progression (Figure 7G). Concurrent ET further suppresses G1 progression by inhibiting c-Myc-mediated E2F amplification, intensifying the blockade of cell-cycle progression. Additionally, we observed that CDK2i is only effective at suppressing CDK2 activity and the growth of drug-resistant cells when combined with CDK4/6i. Notably, overexpression of cyclin E and A fosters resistance to this drug combination. These insights suggest the advantages of continuing CDK4/6i therapy beyond disease progression, both in vitro and in vivo. Moreover, it aligns with previous reports highlighting the importance of dual targeting CDK4/6 and CDK2 (26–30).

Our data indicate that maintaining both CDK4/6i and ET synergistically decelerates cell-cycle progression in drug-resistant cells by further delaying CDK2 activation kinetics and the G1/S transition without affecting the S and G2 phases. This dual effect stems from CDK4/6i causing suboptimal Rb inactivation while ET suppresses the global transcription amplifier c-Myc, collectively leading to diminished E2F transcriptional activity. As a result, this reduced E2F activity lowers the expression of critical cell-cycle genes, such as cyclin E and A, extending the time needed for CDK2 activation. Given that CDK2 plays an essential role in initiating and advancing DNA replication (56,57), its delayed activation significantly prolongs the G1/S transition. Moreover, CDK2 activation also contributes to Rb phosphorylation and inactivation. High CDK2 activity is required to phosphorylate Rb, and CDK2-mediated Rb phosphorylation is tightly coupled with DNA replication timing (7,40). Thus, upon Rb phosphorylation by CDK2 at the G1/S transition, drug-resistant cells may effectively proceed through the cell cycle even under continued CDK4/6i treatment.

Clinical trials evaluating the efficacy of sustained CDK4/6i therapy predominantly use PFS as the primary endpoint (18–22). However, our findings suggest that drug-resistant tumors continue to proliferate despite CDK4/6i maintenance. Consequently, maintaining CDK4/6i appears to slow tumor growth rather than completely arrest it. This underscores the need for clinical trials to consider overall survival and tumor progression rates as more appropriate endpoints for assessing the true benefits of sustained CDK4/6i therapy. Furthermore, the distinct polypharmacology profiles among CDK4/6i (58), with ribociclib being the most specific and abemaciclib the least, may explain the varying therapeutic outcomes observed among these inhibitors (20,59).

Maintaining CDK4/6i treatment post-disease progression may be particularly beneficial for approximately 70% of patients who do not acquire new genetic mutations (60). However, it is important to recognize that resistance to CDK4/6i often arises from mutations in genes associated with mitogenic or hormone-signaling pathways (60–64). These include mutations in PIK3CA, ESR1, FGFR1–3, and HER2, which have been linked to increased c-Myc expression (43,65,66). Additionally, recent studies have identified FAT1 mutations as a driver of CDK4/6i resistance (35,67). These resistance mutations may reduce the efficacy of maintaining CDK4/6i and ET therapy. Moreover, approximately 4.7% of HR+/HER2- breast cancer patients exhibit Rb mutations (60,62), making CDK4/6i treatment unlikely to be effective, thus making its continuation unadvisable in these cases.

In conclusion, our study offers significant mechanistic insights that support the continued use of CDK4/6i and ET in patients with HR+/HER2- breast cancer after disease progression. Our findings also emphasize the necessity of combining CDK2i with CDK4/6i to enhance therapeutic efficacy, reinforcing conclusions from previous studies (26–29,44). Moreover, we underscore the critical role of cyclin E overexpression in driving resistance to the combination of CDK4/6i and CDK2i, which may inform future strategies to overcome therapeutic resistance in breast cancer.

Validation of drug resistance and visualization of cell-cycle progression.

(A) Dose-response curves of breast cancer cells treated with palbociclib (10 nM - 10 µM) for 48 hr. Data are shown as mean ± SEM (n = 3 biological replicates). Solid lines represent sigmoidal best-fit curves. (B-E) Heatmaps of single-cell traces for CDK4/6 (left) and CDK2 (middle) activities, and Cdt1-degron intensity (right) in various conditions: drug-naïve MCF-7 (B) and MDA-MB-231 (C) cells, and drug-resistant MDA-MB-231 cells without (D) or with (E) continuous palbociclib (1µM) treatment. Proliferating cells were identified based on CDK2 activity (>1 for more than 2 hr between 30 and 48 hr).

Slow cell-cycle progression in drug-resistant cells continuously treated with CDK4/6i.

(A-C) Single-cell traces of CDK4/6 (left) and CDK2 (middle) activities and Cdt1-degron intensity (right) aligned by mitosis in drug-naïve MCF-7 (A) and MDA-MB-231 (B) cells and drug-resistant MDA-MB-231 cells without CDK4/6i treatment (C). The time of mitosis is marked in gray. (D-E) Single-cell trace of CDK4/6 activity corresponding to Figure 1G (D) and 1H (E). (F) Single-cell traces of CDK4/6 (left) and CDK2 (middle) activities and Cdt1-degron intensity (right) aligned by mitosis in drug-resistant MDA-MB-231 cells with continuous palbociclib (1 µM) treatment.

Potential non-canonical role of CDK6 in promoting CDK4/6i resistance.

(A-B) Immunoblot showing CDK6 and β-actin expression in drug-naïve and palbociclib-resistant cells (A) and WT and CDK6-KO cells (B). (C-E) Heatmaps of single-cell traces for CDK4/6 (left) and CDK2 (middle) activities, and Cdt1-degron intensity (right) in drug-naïve (C) and drug-resistant WT (D) and CDK6-KO (E) cells with continued palbociclib (1 µM) treatment. Proliferating cells were identified based on CDK2 activity (>1 for more than 2 hr between 30 and 48 hr). (F) Averaged traces of CDK4/6 (left) and CDK2 (right) activities. Data are shown as mean ± 95% confidence intervals (n > 7,000 cells/condition).

Incomplete Rb loss mediates the extended G1-phase progression.

(A) Immunoblot showing total Rb and β-actin in WT, Rb-KO, and palbociclib-resistant cells. (B) G1-phase duration in drug-resistant WT and Rb-KO cells with continued palbociclib (1 µM) treatment. Solid and dashed yellow lines represent mean and median, respectively (n > 4,500 cells). (C) Averaged CDK4/6 (left) and CDK2 (right) activities aligned by mitosis in drug-resistant WT and Rb-KO cells with continued palbociclib (1 µM) treatment (n > 7,500 cells).

Slow and heterogeneous G1/S transition in drug-resistant cells maintained with CDK4/6i treatment.

(A) Single-cell traces showing CDK2 activity aligned by mitosis in proliferating (red) and quiescent (blue) drug-naïve MCF-7 cells. Circles indicate the time of fixation and staining (n = 200 cells). (B-C) Scatterplot of EdU intensity against time since mitosis in drug-resistant cells without (B) and with (C) continuous palbociclib (1 µM) treatment. Red dotted line indicates the S-phase threshold (n = 2,000 cells/condition). (D-E) Scatterplot of EdU intensity (D) and E2F1 mRNA levels (E) against CDK2 activity (n = 2,000 cells/condition).

c-Myc overexpression facilitates CDK4/6i resistance by accelerating cell-cycle progression.

(A) Growth curves of drug-resistant cells with treatment discontinuation or continuous palbociclib (1 µM) treatment, without or with c-Myc overexpression. Data are shown as mean ± SD (n = 3 biological replicates). Statistical significance was determined using one-way ANOVA with Tukey’s post-hoc analysis (** P < 0.01, **** P < 0.0001). (B) Intermitotic time of drug-resistant cells without and with c-Myc overexpression. Solid and dashed yellow lines represent mean and median, respectively. (C-D) Heatmaps of single-cell traces for CDK4/6 (left) and CDK2 (middle) activities, and Cdt1-degron intensity (right) in drug-resistant cells undergoing continuous palbociclib (1 µM) treatment without (C) and with (D) c-Myc overexpression. Proliferating cells were identified based on CDK2 activity (>1 for more than 2 hr between 30 and 48 hr).

Maintaining CDK4/6i and ET synergistically suppresses the growth of drug-resistant cells.

(A) Growth curves of MCF-7 (left) and CAMA-1 (right) cells resistant to palbociclib and fulvestrant withdrawn from drug treatment. Data are shown as the mean ± SD (n = 3 biological replicates). (B) Cell numbers 20 days after treatment with DMSO, palbociclib (1 µM), fulvestrant (500 nM), or their combination. Data are shown as mean ± SD (n = 3 biological replicates). Statistical significance was determined with an unpaired t-test (** P < 0.01, *** P < 0.001, **** P < 0.0001). (C) Heatmaps of single-cell traces for CDK4/6 (left) and CDK2 (middle) activities, and Cdt1-degron intensity (right) in combination drug-resistant cells maintained with palbociclib (1 µM). Proliferating cells were identified based on CDK2 activity (>1 for more than 2 hr between 30 and 48 hr). (D) Single-cell traces of CDK2 activity aligned by mitosis in combination drug-resistant cells treated with continuous fulvestrant (500 nM) (left), palbociclib (1 µM) (middle), or their combination (right). Based on CDK2 activity, cells were classified into proliferation (red) or quiescence (blue). The time of mitosis is marked in gray. (E) Single-cell traces of Cdt1-degron intensity aligned by mitosis in combination drug-resistant cells treated with continuous fulvestrant (500 nM) alone. (F-G) Violin plots showing S-phase duration (n > 220 cells/condition) (F) and G2/M-phase duration (n > 23 cells/condition) (G) in MCF-7 (left) and CAMA-1 (right) cells. Solid and dashed yellow lines indicate mean and median, respectively.

Combining CDK2i with CDK4/6i and ET effectively suppresses the growth of drug-resistant cells.

(A-D) Heatmaps of single-cell traces for CDK4/6 (left) and CDK2 (middle) activities, and Cdt1-degron intensity (right) in drug-resistant cells under various conditions: treatment discontinuation (A), fulvestrant (500 nM) alone (B), fulvestrant + palbociclib (1 µM) (C), or fulvestrant + INX-315 (100 nM) (D) over one week prior to imaging. Proliferating cells were identified based on CDK2 activity (>1 for more than 2 hr between 30 and 48 hr). (E) Violin plots showing S-phase duration (n > 400 cells/condition) (left) and G2/M-phase duration (n > 150 cells/condition) (right) in drug-resistant MCF-7 cells treated with the indicated drug. Solid and dashed yellow lines indicate mean and median, respectively. (F) GSEA plots for hallmark E2F (left) and MYC (right) target genes in palbociclib/fulvestrant-resistant MCF-7 cells treated with the triple combination, compared to other drug conditions.

Cyclin E or A overexpression attenuates full CDK2 inhibition by CDK4/6i and CDK2i combination.

(A) Immunoblot showing the expression of GAPDH with cyclin A (left) or cyclin E (right) in MCF-7 cells with or without cyclin E1 or A2 overexpression. Cells were treated with doxycycline (500 nM) for 24 hr. (B-G) Averaged traces of CDK4/6 (left) and CDK2 (right) activities in MCF-7 cells with or without cyclin E1 and A2 overexpression. Cells were treated with indicated doxycycline (500 nM) 6 hr before the addition of indicated drugs. Data are shown as mean ± 95% confidence intervals (n > 850 cells/condition).

Cyclin E overexpression facilitates adaptation to CDK4/6i and CDK2i combination.

Heatmaps of single-cell traces for CDK4/6 (left) and CDK2 (middle) activities, and Geminin-degron intensity (right) in drug-naïve MCF-7 cells overexpressing cyclin A and treated with palbociclib (1 µM) and INX-315 (100 nM). Proliferating cells were identified based on CDK2 activity (>1 for more than 2 hr between 30 and 48 hr).

Materials and Methods

Cell culture

MCF-7 (ATCC, HTB-22), CAMA1 (ATCC, HTB-21), and MDA-MB-231 (ATCC, CRM-HTB-26) cells were cultured in Dulbecco’s Modified Eagle Medium (DMEM; Genesee Scientific, 25-500) supplemented with 10% Fetal Bovine Serum (FBS; Gibco, A3160601). T47D (ATCC, HTB-133) cells were cultured in RPMI-1640 (Genesee Scientific, 25-206) with 10% FBS. All cell lines were incubated at 37 °C in a humidified atmosphere with 5% CO2. All cell lines were routinely tested and confirmed to be mycoplasma contamination-free.

Drugs and Chemicals

Palbociclib (Selleck Chemicals, S1116), INX-315 (MedchemExpress, HY-162001), and fulvestrant (Selleck Chemicals, S1191) were dissolved in DMSO (Sigma-Aldrich, D2438) for in vitro experiments. For in vivo experiments, palbociclib (MedChemExpress, HY-50767), ribociclib (MedChemExpress, HY-15777), and abemaciclib (MedChemExpress, HY-16297A) were prepared in a solution of corn oil (Spectrum Chemical MFG, CO136) with 10% DMSO. Additional reagents, including 5-Ethynyl-2’-Deoxyuridine (EdU; Sigma-Aldrich, 900584) and AFDye-647 picolyl azide (Click Chemistry Tools, 1300) were used to measure cells undergoing DNA replication.

Antibodies

The following antibodies were employed from Cell Signaling Technology: Rb (9309), Phosphorylated Rb (Ser807/811) (8516), and c-Myc (5605). Additional antibodies were obtained from Abcam for CDK6 (ab124821), Cyclin E (ab32103), and cyclin A (ab32386), and from BD Biosciences for Rb (554136). BD Biosciences Rb antibody was used to visualize Rb protein in cells expressing CDK4/6 sensor. Secondary antibodies from Thermo Scientific included Alexa Fluor 488 goat anti-mouse (A32723) and Alexa Fluor 568 goat anti-rabbit (A11036). For immunoblotting, we used secondary antibodies from LI-COR Biosciences: goat anti-mouse IR Dye 800CW (926-32210) and goat anti-rabbit IR Dye 680RD (926-68071).

DNA constructs and cell line generation

All DNA constructs used in this study were generated as described previously (7,31,38,39,68). Briefly, using Gibson assembly, constructs including H2B-iRFP670-p2a-mCerulean-Cdt1 (a.a.1–100) (Addgene, 223965), H2B-iRFP670-p2a-mCerulean-Geminin (a.a.1–110) (Addgene, 223959), and DHB (a.a.995–1087)-mVenus-p2a-mCherry-Rb (a.a.886–928) (Addgene, 126679) were inserted into pLenti-IRES vectors encoded with puromycin, blasticidin, or neomycin selection markers. Doxycycline-inducible c-Myc, cyclin E1, and cyclin A2 were created by inserting PCR products into the lentiviral pCW57.1 vector (Addgene, 50661) following NheI and BamHI restriction digestion. Stable cell lines were established through lentiviral transduction. For virus production, lentiviral plasmids encoding the genes of interest were transfected into HEK-293T cells along with packaging plasmids pMDLg/pRRE (Addgene, 12251), pRSV-Rev (Addgene, 12253), and envelope plasmid pCMV-VSV-G (Addgene, 8454), using polyethyleneimine transfection reagent in Opti-MEM (Thermo Scientific, 31985070). Following 72 and 96 hr of incubation, the viral supernatant was collected, pooled, and centrifuged at 1,200 rpm for 5 minutes, followed by filtration through a 0.45 µm filter (Millipore, SLHA033SB). The filtered supernatant was then concentrated using an ultra-centrifugal filter (Millipore, UFC910024) at 4,000 rpm for 10 minutes. These concentrated viruses were stored at −80°C until further use. Cells were infected with virus in the presence of 5 µM polybrene for two days and subsequently selected either through antibiotic selection with 1 µg/ml puromycin (InvivoGen, ant-pr), 10 µg/ml blasticidin (InvivoGen, ant-bl), or 800 µg/ml neomycin (Thermo Scientific, BP673-5), or alternatively by sorting cells based on their introduced fluorescent protein using an Influx cell sorter.

CDK6 and Rb knockout cell lines

Rb knockout in MCF-7 cells was previously described and validated (38). To establish CDK6 knockout, MCF-7 cells were transfected with separate crRNA:tracrRNA-ATTO550 plus CRISPR-Cas9 ribonucleoprotein complexes. Successfully transfected cells, identified by ATTO550 fluorescence, were isolated into single cells by fluorescence-activated cell sorting and then grown into colonies. Knockouts were confirmed by immunoblot. Two pooled CRISPR-Cas9 guides were used to knockout CDK6 gene:

CDK6 gRNA 1: 5’-GACCACGUUGGGGUGCUCGAGUUUUAGAGCUAUGCU-3’

CDK6 gRNA 2: 5’-CUGGACUGGAGCAAGACUUCGUUUUAGAGCUAUGCU-3’

Drug-resistant cell lines

To induce resistance to CDK4/6i, breast cancer cells were subjected to chronic exposure to palbociclib (1 µM) or a combination of palbociclib (1 µM) and fulvestrant (500 nM) over 1–2 months. The cell culture media containing the drugs was refreshed every 2–3 days throughout the experimental period. Resistance development was confirmed by assessing drug sensitivity through titration of various drug concentrations and calculating the half-maximal inhibitory concentration (IC50) based on the proportion of S-phase cells compared to drug-naïve controls.

Live-cell reporters

Histone 2B (H2B) tagged with iRFP670 was used to define nuclear regions and track individual cells. Live-cell reporters were employed to monitor CDK activity and cell-cycle transitions. Fluorescently tagged Cdt1 degron (a.a.1–100) (33) was used to detect cell-cycle phase transitions. Fluorescent geminin protein was used to track S-phase entry. CDK4/6 and CDK2 kinase translocation reporters (KTRs) (31,32) were used to monitor the phosphorylation state of specific substrates. The CDK4/6 substrate was the C-terminal domain of Rb (a.a.886–928), and the CDK2 substrate was DNA helicase B (a.a.994–1087). Fluorescence translocation between the nucleus and cytoplasm provided a quantitative readout of CDK activity. The cytoplasmic/nuclear fluorescence intensity ratio was used to represent kinase activity. Since the CDK4/6 KTR contains a degenerate CDK2 substrate motif from the CDK2 KTR and thus detects activities measured by the CDK2 sensor during the S and G2 phases, we applied a correction factor derived from a linear regression based on the CDK2 sensor, as described previously (31,38,68). CDK4/6 activity was adjusted by subtracting a calculated fraction of the CDK2 KTR signal, as follows:

MCF-7: CDK4/6 activity = CDK4/6 reporter activity – (0.41 × CDK2 reporter activity)

MDA-MB-231: CDK4/6 activity = CDK4/6 reporter activity – (0.35 × CDK2 reporter activity)

Immunofluorescence and mRNA fluorescence in situ hybridization (FISH)

Cells were seeded in glass-bottom 96 well plates (Cellvis, P96-1.5H-N) at least one night prior to each experiment. The percentage of S-phase cells was determined using EdU incorporation. Cells were incubated with EdU (10 µM) for 15 minutes at 37°C, followed by fixation with 4% formaldehyde (Thermo Scientific, 28906) containing 10 mM HEPES (Sigma-Aldrich, H3537) in PBS for 15 minutes at room temperature. Permeabilization was then performed using 0.2% Triton X-100 (Sigma-Aldrich, T8787) in PBS (PBS-T) for 15 minutes. Visualization of EdU incorporation was achieved using a click reaction in 2 mM CuSO4 (Sigma-Aldrich, C1297), 20 mg/ml sodium ascorbate (Sigma-Aldrich, A4034), and 3 µM Alexa Fluor Dye 647 picolyl azide in Tris-Buffered Saline (TBS) (Sigma-Aldrich, T6066) (pH 8.3) for 15 minutes. For immunofluorescence staining, cells were incubated in blocking buffer (10% FBS, 1% BSA, 0.1% Triton X-100, and 0.01% NaN3 in PBS) for 1 hour. Cells were then incubated with primary antibodies for Phospho-Rb (Cell Signaling Technology, 8516), Rb (BD Biosciences, 554136), and c-Myc (Cell Signaling Technology, 5605), diluted 1:2000 in blocking buffer, overnight at 4°C. The following day, Alexa Fluor 488- or 568-conjugated secondary antibodies were applied at 1:2000 dilution in blocking buffer for 1 hour at room temperature. DNA was stained with Hoechst 33342 dye (Thermo Scientific, 62249) (10 mg/ml), diluted in PBS (1:10000) for 15 minutes at room temperature prior to imaging. Cells were washed with PBS between each step and stored in PBS prior to imaging. For mRNA FISH, the ViewRNA ISH Cell Assay Kit (Thermo Scientific, QVC0001) was used along with a probe targeting E2F1 (Thermo Scientific, VA1-12108-VC). Following live-cell imaging, cell fixation and permeabilization were conducted as described in the immunofluorescence section. The fluorescent protein signal was then photobleached using 3 % H2O2 (Sigma-Aldrich, H1009) with 20 mM HCl (Sigma-Aldrich, 320331) in PBS for 2 hours at room temperature. Probe hybridization, amplification, and Alexa Fluor 555 labeling were carried out following the manufacturer’s protocols.

Growth curve measurement

To assess overall cell proliferation, 5,000 cells were seeded into each well of a 24-well plate (Fisher Scientific, FB012929). Cells were harvested at the indicated time points by trypsinization, resuspended in 1 ml growth media, and manually counted using a hemocytometer (Hausser Scientific, HS-3510). Cells were then replated as needed for subsequent time points.

Live and Fixed cell and Tumor tissue image acquisition

Cells were seeded and maintained at 30–80% confluency during experiments in 96-well glass-like polymer bottom plates (Cellvis, P96-1.5P). Phenol red-free media was used to minimize background fluorescence. Cell imaging was conducted using a microscope mounted onto an inverted Eclipse Ti-2 body (Nikon), and multichannel fluorescent images were captured with a Hamamatsu ORCA-Fusion camera. Acquisition used either a 10× (Nikon CFI Plan Apo Lambda, 0.45 NA, no binning) or a 20× (Nikon CFI Plan Apo Lambda, 0.75 NA, 2×2 binning) objective. Live-cell imaging was performed in a 37 °C humidified chamber with 5% CO2, capturing 3 sites per well at 12-min intervals. 9 (10× objective) or 32 (20× objective) sites per well were collected for fixed-cell imaging. Total light exposure time was kept under 500 ms for each time point. mRNA FISH images were acquired with 1.5 µm z-stack intervals. Whole tumor cross-sections were imaged using an Axio Observer 7 microscope with Apotome 2 (Zeiss) and a Hamamatsu ORCA-Flash 4 camera, using a 10× (Zeiss Plan-Apochromat, 0.45 NA, no binning) objective at a 1.5 µm z-stack interval.

Image processing and analysis

Live- and fixed-cell image processing and analysis were conducted using custom MATLAB scripts (MathWorks, R2021a). To reduce biased light illumination, flat-field correction was applied. In fixed-cell analysis, nuclei were segmented based on the Hoechst signal according to a threshold of histogram curvature. For live-cell analysis, cell segmentation was performed using the H2B-iRFP670 signal and a Laplacian of Gaussian blob detector. A marker-based watershed algorithm was then employed to separate neighboring nuclei. To correct for background signal, the 50th percentile of pixels in the remaining background was subtracted after masking nuclei. In the analysis transitioning from live to fixed cells, cells were initially segmented and then aligned between the fixed cell data and the last frame of the live-cell experiment. A deflection-bridging algorithm was utilized to track cell movement over time during live-cell acquisition. Mitotic events were identified by isolating two daughter cell nuclei adjacent to a prior nucleus, with H2B signals approximately 45–55% of the mother cell’s nuclear signal. In live-cell analysis, cytoplasmic signal determination involved measuring the median pixel intensity within a ring positioned between 0.65–3.25 µm outside the nuclear perimeter. Cells with overlapping cytoplasmic rings were excluded from the final analysis. For RNA FISH, the whole cell area was estimated by creating a mask including and extending 50 µm from the nucleus. Overlapping cells with adjacent cell nuclei were excluded from the analysis. Raw images underwent a top hat filter with a 4 µm radius circular kernel to create the FISH puncta mask. Tissue image analysis was performed using ImageJ software.

Immunoblot

Cells were rinsed with ice cold PBS before incubation in 300 µl CHAPS lysis buffer (50 mM Tris-HCl, 150 mM NaCl, 1 mM EDTA, 10 mM NEM, 0.3% CHAPS, 1 mM PMSF) supplemented with 1× Halt Protease inhibitor (Thermo Scientific, 1861279) and 1× Pierce phosphatase inhibitor (Roche, 4906845001) for 5 minutes on ice. Lysates were collected and then spun at 4°C for 10 minutes at 14,000 RCF, and supernatants were used for subsequent experiments. Protein concentration was determined by spectrophotometric interpolation after standard curve construction using the Pierce 660 nm protein assay reagent (Thermo Scientific, 22660) according to the manufacturer’s instructions. 12 µg of protein were incubated in LDS sample buffer (Thermo Scientific, NP0007) at 70°C for 10 minutes before separation on NuPAGE 4–12% Bis-Tris gel (Thermo Scientific, NP0322BOX). Protein was then transferred to a 0.45 µm mini PVDF transfer membrane (Bio-Rad, 1620174) using a Bio-Rad Trans-Blot Turbo system (Bio-Rad, 1704150) and 1× Trans-Blot Transfer Buffer (Bio-Rad, 10026938). Membranes were then blocked with Intercept Blocking Buffer (LI-COR, 927-60001) at room temperature for 1 hour. Immunoblots were subsequently incubated with either Rb (Cell Signaling Technology, 9309), Phospho-Rb (Cell Signaling Technology, 8516), CDK6 (Abcam, ab124821) or Beta-actin (Cell Signaling Technology, 3700S) primary antibody diluted 1:1000 in Intercept Blocking Buffer overnight at 4°C. After primary incubation, membranes were washed with TBS-T (20 mM Tris, 150 mM NaCl, 0.1% Tween 20, pH 7.5) and incubated in either goat-anti mouse IR Dye 800CW (LI-COR, 926-32210) or goat-anti rabbit IR Dye 680RD (LI-COR, 926-68071) at 1:2000 in Intercept Blocking Buffer for 2 hours at room temperature. After rinsing with TBS-T, membranes were visualized using the LI-COR Odyssey Infrared Imaging System. Immunoblot images were processed with ImageStudio Lite Ver5.2.

RNAseq

RNA isolation was performed using the GeneJET RNA Purification Kit (Thermo Scientific, K0732) according to manufacturer’s instruction. After trypsinization, cells were rinsed with PBS then lysed in suspension using 600 µL of the supplied lysis buffer supplemented with 14.3M β-mercaptoethanol (Bio-Rad, 1610710). After homogenization and addition of 360 µL 100% ethanol (Fisher Scientific, BP28184), RNA was isolated and collected by centrifugation before final elution with DNase/RNase-free distilled water (Thermo Scientific, 10977-015). RNA sequencing was then performed by Azenta Life Sciences. Statistical analysis of count matrix was performed using DESeq2 R package (v1.44.0) (69). Log2 fold-change shrinkage was applied to improve effect size estimates using the ‘apeglm’ method (70). Gene sets were obtained from the Molecular Signatures Database (Hallmark, C2, C5) utilizing the msigdbr package (version 7.5.1). Heatmaps were created with the ComplexHeatmap package (71). Gene set enrichment analysis was conducted with hallmark annotations from MSigDB using the R package clusterProfiler (v4.12.6). Ranked gene lists were input with a maximum gene set size of 500, and significance was evaluated through 10,000 permutations. Terms were considered significantly enriched if their P value was below 0.05 and their False Discovery Rate (FDR) was under 0.25.

Animal

All experiments were conducted in accordance with NIH guidelines and protocols approved by the Institutional Animal Care and Use Committee (IACUC) at Columbia University Irving Medical Center. Six-week-old female Foxn1-/-J:NU mice (The Jackson Laboratory, 007850) were housed in a barrier facility with ad libitum access to food and water on a 12-hour light-dark cycle. Mice were regularly monitored in compliance with institutional guidelines for ethical endpoints.

Xenograft experiments

Female J:NU mice were anesthetized with ketamine (45 mg/kg) plus xylazine (5 mg/kg) by intraperitoneal injection. Mice were then orthotopically injected with drug-naïve MCF-7 cells (8×106 cells per mouse) suspended in 1:1 ratio of PBS and Geltrex LDEV-Free Reduced Growth Factor Basement Membrane Matrix (Thermo Scientific, A1413201) directly into the abdominal mammary fat pad. Tumors were measured biweekly using a digital caliper (volume = [width2 × length] × 1/2). Once tumor volumes reached ~100 mm3, mice began daily treatment with palbociclib (50 mg/kg) by oral gavage. When tumors reached between 145–155 mm3, mice were randomly assigned to 1) discontinuation of treatment, 2) daily palbociclib (50 mg/kg) treatment, 3) daily abemaciclib (50 mg/kg) treatment, or 4) daily ribociclib (200 mg/kg) treatment for 32 days. Upon treatment completion, mice were anesthetized by ketamine/xylazine, then perfused with 1% formaldehyde before tumor extraction.

Immunohistochemistry

After vascular perfusion with 1% formaldehyde in PBS, tumors were fixed in 1% formaldehyde at 4°C for 1 hour. Following fixation, tissues were washed with PBS-T and incubated in 30% sucrose at 4°C overnight. Samples were embedded in optimal cutting temperature compound (Fisher Scientific, 23-730-571), frozen at −80°C, and sectioned into 40 µm sections onto Superfrost Plus microscope slides (Fisher Scientific, 12-550-15). Tumor sections were rehydrated in PBS-T and subsequently blocked with 5% normal goat serum (The Jackson Laboratories, 005-000-121) in PBS-T for 1 hour at room temperature. The sections were then incubated with recombinant anti-Rb monoclonal antibody (Abcam, ab181616) in 5% normal goat serum in PBS-T at 1:1000 overnight at room temperature. Following washing with PBS-T, the sections were incubated with Alexa Fluor 488 goat anti-rabbit (Thermo Scientific, A32731) at 1:500 in PBS-T at room temperature for 4 hours. DNA was counterstained with DAPI (Sigma-Aldrich, D9542) 1:500 in PBS-T for 10 minutes, then mounted with Fluoromount-G (Invitrogen, 00-4958-02) for image acquisition.

Statistics and Reproducibility

Statistical analyses were performed using GraphPad Prism 10.2.0. For parametric data, we utilized unpaired two-tailed Student’s t-tests or one-way ANOVA with Tukey’s post hoc tests, as appropriate, to assess pairwise differences. When comparing two-paired groups, paired two-tailed Student’s t-tests were employed. Statistical test results are reported in the figure legends and detailed in Supplementary Table 1. All experiments were conducted in at least two independent experiments.

eLife Assessment

This study presents fundamental insights into overcoming resistance in hormone receptor-positive breast cancer by demonstrating that sustained CDK4/6 inhibitor treatment, either alone or in combination with CDK2 inhibitors, significantly suppresses the growth of drug-resistant cancer cells. The findings are supported by compelling evidence from both in vitro cell line experiments and in vivo mouse models, highlighting the therapeutic potential of maintaining CDK4/6 inhibitors beyond disease progression. Additionally, the identification of cyclin E overexpression as a key driver of resistance offers a target that will be of value for future therapeutic strategies, potentially improving outcomes for patients with advanced breast cancer.

Reviewer #1 (Public review):

Summary:

In this manuscript, the authors identified that

(1) CDK4/6i treatment attenuates the growth of drug-resistant cells by prolongation of the G1 phase;

(2) CDK4/6i treatment results in an ineffective Rb inactivation pathway and suppresses the growth of drug-resistant tumors;

(3) Addition of endocrine therapy augments the efficacy of CDK4/6i maintenance;

(4) Addition of CDK2i with CDK4/6 treatment as second-line treatment can suppress the growth of resistant cell;

(5) The role of cyclin E as a key driver of resistance to CDK4/6 and CDK2 inhibition.

Strengths:

To prove their complicated proposal, the authors employed orchestration of several kinds of live cell markers, timed in situ hybridization, IF and Immunoblotting. The authors strongly recognize the resistance of CDK4/6 + ET therapy and demonstrated how to overcome it.

Weaknesses:

The authors need to underscore their proposed results from what is to be achieved by them and by other researchers.

Reviewer #2 (Public review):

Summary:

This study elucidated the mechanism underlying drug resistance induced by CDK4/6i as a single agent and proposed a novel and efficacious second-line therapeutic strategy. It highlighted the potential of combining CDK2i with CDK4/6i for the treatment of HR+/HER2- breast cancer.

Strengths:

The study demonstrated that CDK4/6 induces drug resistance by impairing Rb activation, which results in diminished E2F activity and a delay in G1 phase progression. It suggests that the synergistic use of CDK2i and CDK4/6i may represent a promising second-line treatment approach. Addressing critical clinical challenges, this study holds substantial practical implications.

Weaknesses:

(1) Drug-resistant cell lines: Was a drug concentration gradient treatment employed to establish drug-resistant cell lines? If affirmative, this methodology should be detailed in the materials and methods section.

(2) What rationale informed the selection of MCF-7 cells for the generation of CDK6 knockout cell lines? Supplementary Figure 3. A indicates that CDK6 expression levels in MCF-7 cells are not notably elevated.

(3) For each experiment, particularly those involving mice, the author must specify the number of individuals utilized and the number of replicates conducted, as detailed in the materials and methods section.

(4) Could this treatment approach be extended to triple-negative breast cancer?

Reviewer #3 (Public review):

Summary:

In their manuscript, Armand and colleagues investigate the potential of continuing CDK4/6 inhibitors or combining them with CDK2 inhibitors in the treatment of breast cancer that has developed resistance to initial therapy. Utilizing cellular and animal models, the research examines whether maintaining CDK4/6 inhibition or adding CDK2 inhibitors can effectively control tumor growth after resistance has set in. The key findings from the study indicate that the sustained use of CDK4/6 inhibitors can slow down the proliferation of cancer cells that have become resistant, and the combination of CDK2 inhibitors with CDK4/6 inhibitors can further enhance the suppression of tumor growth. Additionally, the study identifies that high levels of Cyclin E play a significant role in resistance to the combined therapy. These results suggest that continuing CDK4/6 inhibitors along with the strategic use of CDK2 inhibitors could be an effective strategy to overcome treatment resistance in hormone receptor-positive breast cancer.

Strengths:

(1) Continuous CDK4/6 Inhibitor Treatment Significantly Suppresses the Growth of Drug-Resistant HR+ Breast Cancer: The study demonstrates that the continued use of CDK4/6 inhibitors, even after disease progression, can significantly inhibit the growth of drug-resistant breast cancer.

(2) Potential of Combined Use of CDK2 Inhibitors with CDK4/6 Inhibitors: The research highlights the potential of combining CDK2 inhibitors with CDK4/6 inhibitors to effectively suppress CDK2 activity and overcome drug resistance.

(3) Discovery of Cyclin E Overexpression as a Key Driver: The study identifies overexpression of cyclin E as a key driver of resistance to the combination of CDK4/6 and CDK2 inhibitors, providing insights for future cancer treatments.

(4) Consistency of In Vitro and In Vivo Experimental Results: The study obtained supportive results from both in vitro cell experiments and in vivo tumor models, enhancing the reliability of the research.

(5) Validation with Multiple Cell Lines: The research utilized multiple HR+/HER2- breast cancer cell lines (such as MCF-7, T47D, CAMA-1) and triple-negative breast cancer cell lines (such as MDA-MB-231), validating the broad applicability of the results.

Weaknesses:

(1) The manuscript presents intriguing findings on the sustained use of CDK4/6 inhibitors and the potential incorporation of CDK2 inhibitors in breast cancer treatment. However, I would appreciate a more detailed discussion of how these findings could be translated into clinical practice, particularly regarding the management of patients with drug-resistant breast cancer.

(2) While the emergence of resistance is acknowledged, the manuscript could benefit from a deeper exploration of the molecular mechanisms underlying resistance development. A more thorough understanding of how CDK2 inhibitors may overcome this resistance would be valuable.

(3) The manuscript supports the continued use of CDK4/6 inhibitors, but it lacks a discussion on the long-term efficacy and safety of this approach. Additional studies or data to support the safety profile of prolonged CDK4/6 inhibitor use would strengthen the manuscript.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public review):

Summary:

In this manuscript, the authors identified that

(1) CDK4/6i treatment attenuates the growth of drug-resistant cells by prolongation of the G1 phase;

(2) CDK4/6i treatment results in an ineffective Rb inactivation pathway and suppresses the growth of drugresistant tumors;

(3) Addition of endocrine therapy augments the efficacy of CDK4/6i maintenance;

(4) Addition of CDK2i with CDK4/6 treatment as second-line treatment can suppress the growth of resistant cell;

(5) The role of cyclin E as a key driver of resistance to CDK4/6 and CDK2 inhibition.

Strengths:

To prove their complicated proposal, the authors employed orchestration of several kinds of live cell markers, timed in situ hybridization, IF and Immunoblotting. The authors strongly recognize the resistance of CDK4/6 + ET therapy and demonstrated how to overcome it.

Weaknesses:

The authors need to underscore their proposed results from what is to be achieved by them and by other researchers.

Reviewer #2 (Public review):

Summary:

This study elucidated the mechanism underlying drug resistance induced by CDK4/6i as a single agent and proposed a novel and efficacious second-line therapeutic strategy. It highlighted the potential of combining CDK2i with CDK4/6i for the treatment of HR+/HER2- breast cancer.

Strengths:

The study demonstrated that CDK4/6 induces drug resistance by impairing Rb activation, which results in diminished E2F activity and a delay in G1 phase progression. It suggests that the synergistic use of CDK2i and CDK4/6i may represent a promising second-line treatment approach. Addressing critical clinical challenges, this study holds substantial practical implications.

Weaknesses:

(1) Drug-resistant cell lines: Was a drug concentration gradient treatment employed to establish drug-resistant cell lines? If affirmative, this methodology should be detailed in the materials and methods section.

We greatly appreciate the reviewer for raising this important question. In the revised manuscript, we have updated the methods section (“Drug-resistant cell lines”) to more precisely describe how the drug-resistant cell lines were established.

(2) What rationale informed the selection of MCF-7 cells for the generation of CDK6 knockout cell lines? Supplementary Figure 3. A indicates that CDK6 expression levels in MCF-7 cells are not notably elevated.

We appreciate the reviewer’s insightful question about the rationale for selecting MCF-7 cells to generate CDK6 knockout cell lines. This choice was guided by prior studies highlighting the significant role of CDK6 in mediating resistance to CDK4/6 inhibitors (21-24). Moreover, we observed a 4.6-fold increase in CDK6 expression in CDK4/6i resistant MCF-7 cells compared to their drug-naïve counterparts (Supplementary Figure 3A). While we did not detect notable differences in CDK4/6 activity between wild-type and CDK6 knockout cells under CDK4/6 inhibitor treatment, these findings point to a potential non-canonical function of CDK6 in conferring resistance to CDK4/6 inhibitors.

(3) For each experiment, particularly those involving mice, the author must specify the number of individuals utilized and the number of replicates conducted, as detailed in the materials and methods section.

We sincerely thank the reviewer for bringing this to our attention. In the revised manuscript, we have explicitly stated the number of replicates and mice used for each experiment as appropriate in figure legends and relevant text to ensure transparency and clarity.

(4) Could this treatment approach be extended to triple-negative breast cancer?

We greatly appreciate the reviewer’s inquiry about extending our findings to triple-negative breast cancer (TNBC). Based on the data presented in Figure 1 and Supplementary Figure 2, which include the TNBC cell line MDA-MB-231, we expect that the benefits of maintaining CDK4/6 inhibitors could indeed be applicable to TNBC with an intact Rb/E2F pathway. Additionally, our recent paper (25) indicates a similar mechanism in TNBC.

Reviewer #3 (Public review):

Summary:

In their manuscript, Armand and colleagues investigate the potential of continuing CDK4/6 inhibitors or combining them with CDK2 inhibitors in the treatment of breast cancer that has developed resistance to initial therapy. Utilizing cellular and animal models, the research examines whether maintaining CDK4/6 inhibition or adding CDK2 inhibitors can effectively control tumor growth after resistance has set in. The key findings from the study indicate that the sustained use of CDK4/6 inhibitors can slow down the proliferation of cancer cells that have become resistant, and the combination of CDK2 inhibitors with CDK4/6 inhibitors can further enhance the suppression of tumor growth. Additionally, the study identifies that high levels of Cyclin E play a significant role in resistance to the combined therapy. These results suggest that continuing CDK4/6 inhibitors along with the strategic use of CDK2 inhibitors could be an effective strategy to overcome treatment resistance in hormone receptor-positive breast cancer.

Strengths:

(1) Continuous CDK4/6 Inhibitor Treatment Significantly Suppresses the Growth of Drug-Resistant HR+ Breast Cancer: The study demonstrates that the continued use of CDK4/6 inhibitors, even after disease progression, can significantly inhibit the growth of drug-resistant breast cancer.

(2) Potential of Combined Use of CDK2 Inhibitors with CDK4/6 Inhibitors: The research highlights the potential of combining CDK2 inhibitors with CDK4/6 inhibitors to effectively suppress CDK2 activity and overcome drug resistance.

(3) Discovery of Cyclin E Overexpression as a Key Driver: The study identifies overexpression of cyclin E as a key driver of resistance to the combination of CDK4/6 and CDK2 inhibitors, providing insights for future cancer treatments.

(4) Consistency of In Vitro and In Vivo Experimental Results: The study obtained supportive results from both in vitro cell experiments and in vivo tumor models, enhancing the reliability of the research.

(5) Validation with Multiple Cell Lines: The research utilized multiple HR+/HER2- breast cancer cell lines (such as MCF-7, T47D, CAMA-1) and triple-negative breast cancer cell lines (such as MDA-MB-231), validating the broad applicability of the results.

Weaknesses:

(1) The manuscript presents intriguing findings on the sustained use of CDK4/6 inhibitors and the potential incorporation of CDK2 inhibitors in breast cancer treatment. However, I would appreciate a more detailed discussion of how these findings could be translated into clinical practice, particularly regarding the management of patients with drug-resistant breast cancer.

Thank you to the reviewer for this crucial comment. In the revised Discussion, we've broadened our exploration of clinical translation. Specifically, we emphasize that ongoing CDK4/6 inhibition, although not fully stopping resistant tumors, significantly slows their growth and may offer a therapeutic window when combined with ET and CDK2 inhibition. We also note that these approaches may work best for patients without Rb loss or newly acquired resistance-driving mutations, and that cyclin E overexpression could be a biomarker to inform patient selection. These points together highlight that our findings provide a mechanistic understanding and potential framework for clinical trials testing maintenance CDK4/6i with selective addition of CDK2i as a secondline strategy in drug-resistant HR+/HER2- breast cancer.

(2) While the emergence of resistance is acknowledged, the manuscript could benefit from a deeper exploration of the molecular mechanisms underlying resistance development. A more thorough understanding of how CDK2 inhibitors may overcome this resistance would be valuable.

We thank the reviewer for this valuable suggestion. In the revised manuscript, we have expanded our Discussion to more explicitly synthesize the molecular mechanisms of resistance and how CDK2 inhibitors counteract them. Specifically, we describe how sustained CDK4/6 inhibition drives a non-canonical route of Rb degradation, resulting in inefficient E2F activation and prolonged G1 phase progression. We also highlight the role of c-Myc in amplifying E2F activity and promoting resistance, and we show that continued ET mitigates this effect by suppressing c-Myc. Importantly, we demonstrate that CDK2 inhibition alone cannot fully suppress the growth of resistant cells, but when combined with CDK4/6 inhibition, it produces durable repression of E2F and Myc target gene programs and significantly delays the G1/S transition. Finally, we identify cyclin E overexpression as a key mechanism of escape from dual CDK4/6i + CDK2i therapy, suggesting its potential as a biomarker for patient stratification . Together, these findings provide a detailed mechanistic rationale for how CDK2 inhibition can overcome specific pathways of resistance in HR+/HER2- breast cancer.

(3) The manuscript supports the continued use of CDK4/6 inhibitors, but it lacks a discussion on the long-term efficacy and safety of this approach. Additional studies or data to support the safety profile of prolonged CDK4/6 inhibitor use would strengthen the manuscript.

We appreciate the reviewer’s insightful comment. In the revised manuscript, we emphasize the longterm efficacy and safety considerations of sustained CDK4/6 inhibition. Clinical trial and retrospective data have shown that continued CDK4/6i therapy can extend progression-free survival in selected patients, while maintaining a favorable safety profile (26-28). We have updated the Discussion to highlight these findings more explicitly, underscoring that while prolonged CDK4/6 inhibition slows but does not fully arrest tumor growth, it remains a clinically viable strategy when balanced against its manageable toxicity profile.

Reviewer #1 (Recommendations for the authors):

It is well known that the combination therapy of CDK4/6i and ET has therapeutic benefits in ER(+) HER2(-) advanced breast cancer. However, drug resistance is a problem, and second-line therapy to solve this problem has not been established. Although some parts of the research results are already reported, the authors confirmed them by employing live cell markers, and further proved and suggested how to overcome this resistance in detail. This part is considered novel.

Overall, this research manuscript is eligible to be accepted with the appropriate addressing of questions.

(1)The effects and biochemical changes of combination therapy of CDK4/6i and CDK2i are already known in several papers. The author needs to highlight the differences between the author's research and that of otherresearchers.

We thank the reviewer for the opportunity to clarify the novelty of our findings in the context of prior studies on CDK4/6i and CDK2i combination therapy. In the revised manuscript, we have updated the Discussion section to more clearly delineate how our work extends and differs from existing research.

Specifically, we now state:

Page 12: The combination of CDK4/6i and ET has reshaped treatment for HR+/HER2- breast cancer (1-8). However, resistance commonly emerges, and no consensus second-line standard is established. Our data show that continued CDK4/6i treatment in drug-resistant cells engages a non-canonical, proteolysis-driven route of Rb inactivation, yielding attenuated E2F output and a pronounced delay in G1 progression (Figure 7G). Concurrent ET further deepens this blockade by suppressing c-Myc-mediated E2F amplification, thereby prolonging G1 and slowing population growth. Importantly, CDK2 inhibition alone was insufficient to control resistant cells. Robust suppression of CDK2 activity and resistant-cell growth required CDK2i in combination with CDK4/6i, consistent with prior reports supporting dual CDK targeting (9-16). Moreover, cyclin E, and in some contexts cyclin A, blunted the efficacy of the CDK4/6i and CDK2i combination by reactivating CDK2. Together, these findings provide a mechanistic rationale for maintaining CDK4/6i beyond progression and support testing ET plus CDK4/6i with the strategic addition of CDK2i, as evidenced by concordant in vitro and in vivo results.

(2) Regarding Figures 3H and 3I, I wonder if it is live cell imaging results or if the authors counter each signal via timed IF staining slides? If live cell imaging is used, the authors need to present the methods.

We appreciate the reviewer’s question. Figures 3H and 3I derive from a live–fixed correlative pipeline rather than purely live imaging or independently timed IF slides. We first imaged asynchronously proliferating cells live for ≥48 h to (i) segment/track nuclei with H2B fluorescence, (ii) define mitotic exit (t = 0 at anaphase), and (iii) record CDK2 activity using a CDK2 KTR in the last live frame. Immediately after the live acquisition, we pulsed EdU (10 µM, 15 min) and fixed the same wells, photobleached fluorescent proteins (3% H₂O₂ + 20 mM HCl, 2 h, RT) to prevent crosstalk, and then performed click-chemistry EdU detection, IF for phospho-Rb (Ser807/811) and total Rb, and RNA FISH for E2F1. Fixed-cell readouts (p-Rb positivity, EdU incorporation, E2F1 mRNA puncta) were mapped back to each single cell’s live-derived time since mitosis and/or CDK2 activity, enabling the kinetic plots shown in Fig. 3H–I.

To ensure transparency and reproducibility, we added detailed methods describing this workflow in the “Immunofluorescence and mRNA fluorescence in situ hybridization (FISH)” section under a dedicated “live– fixed pipeline” paragraph, and we cross-referenced acquisition and analysis parameters in “Live- and fixed-cell image acquisition” and “Image processing and analysis.” These updates specify: EdU pulse/fix conditions, photobleaching, antibodies/probes, imaging hardware and channels, segmentation/tracking, mitosis alignment, background correction, and how fixed readouts were binned/quantified as functions of time after mitosis and CDK2 activity.

(3) Regarding Figure 3F, seven images were obtained in same fields? The author needs to describe the meaning of the white image and the yellow and blue image of the bottom in detail.

Thank you for raising this point. All seven panels in Fig. 3F are from the same field of view. The top row shows the raw channels (Hoechst, p-Rb, total Rb, and E2F1 RNA FISH). The bottom row shows the corresponding processed outputs from that field: (i) nuclear segmentation, (ii) phosphorylated Rb-status classification, and (iii) cell boundaries used for single-cell RNA-FISH quantification. We have revised the figure legend to make this explicit.

(4) The author showed E2F mRNA by ISH, but in fact, RB does not suppress E2F mRNA but suppresses protein, so the author needs to confirm E2F at the protein level.

We sincerely appreciate the reviewer’s thoughtful suggestion to examine E2F1 at the protein level. In our study, we focused on E2F1 mRNA expression because it is a well-established and biologically meaningful readout of E2F1 transcriptional activity. Due to its autoregulatory nature (17), the release of active E2F1 protein from Rb induces the transcription of E2F1 itself, creating a positive feedback loop. As a result, E2F1 mRNA abundance serves as a direct and reliable proxy for E2F1 protein activity (18-20). Thus, quantifying E2F1 mRNA provides a biologically relevant and mechanistic indicator of Rb-E2F pathway status. To clarify this rationale, we have updated the Results section and added references supporting our use of E2F1 mRNA as a readout for E2F1 activity.

(5) Is it possible to synchronize cells (nocodazole shake-off, Double thymidine block) under the presence of cdk4/6i? If so, then the authors need to demonstrate the delay of G1 progression via immunoblotting.

We thank the reviewer for this constructive suggestion. To address it, we performed nocodazole synchronization followed by release and monitored cell-cycle progression in the presence or absence of CDK4/6 inhibition.

Specifically, we added the following new datasets to the revised manuscript:

Fig. 3L: Live single-cell trajectories of CDK4/6 and CDK2 activities alongside the Cdt1-degron reporter after 14 hours of nocodazole (250 nM) treatment and release. We compared the averaged traces of CDK4/6 and CDK2 activities and Cdt1 intensity in parental cells (gray) and resistant cells with (red) and without (blue) CDK4/6i maintenance. These data show suppressed and delayed CDK2 activation, as well as a right-shifted S-phase entry, particularly under continuous CDK4/6 inhibition.

Fig. 3M: Fixed-cell EdU pulse-labeling at 4, 6, 8, 12, 16, and 24 h post-release further confirms a significant delay in S-phase entry and prolonged G1 duration in CDK4/6i-maintained cells compared with naïve and withdrawn conditions.

Together, these results directly demonstrate the delay in G1 progression following synchronized mitotic exit under CDK4/6 inhibition.

(6) In Figure 5C the authors showed a violin plot of c-Myc level. Is this Immunohistochemical staining? The authors need to clarify the methods.

Thank you for flagging this. The c-Myc measurements in Fig. 5C are from immunofluorescence (IF), not IHC. We now state this explicitly in the legend.

(7) Regarding Live cell immunofluorescence tracing of live-cell reporters, the author needs to clarify the methods (excitation, emission), name of instruments, and software used.

To address this, we have expanded the “Live-cell, fixed-cell, and tumor tissue image acquisition” section in the Materials and Methods.

(8) Lines 475 SF1A, the authors need to correct typos. Naïve Naïve.

We greatly appreciate the reviewer’s attention to this detail and have ensured all typos have been addressed.

(9) The authors need to unify Cdt1-degron(legends) Vs Cdt1 degron (figures).

We greatly appreciate your attention to this discrepancy. Language referring to the Cdt1 degron has been unified between figures and legends.

Reviewer #3 (Recommendations for the authors):

(1) While the manuscript discusses the selection of doses for CDK4/6 inhibitors and CDK2 inhibitors, there is a lack of detailed data on the dose-response relationship. Additional data on the effects of different doses would be beneficial.

We appreciate the reviewer’s important comment. To address it, we performed additional dose– response experiments testing a range of CDK4/6i and CDK2i concentrations. These analyses revealed a clear synergistic interaction between the two inhibitors. The new data are now presented in Figure 6G and Supplementary Figure 8F of the revised manuscript.

(2) In clinical trials, the criteria for patient selection are crucial for interpreting study outcomes. A detailed description of the patient selection criteria should be provided.

We thank the reviewer for bringing this important point to our attention. In the revised manuscript, we have clarified the patient selection criteria relevant to the interpretation of clinical outcomes. Specifically, we note that retrospective analyses suggest patients with indolent disease and no prior chemotherapy may benefit most from continued CDK4/6i plus ET. Moreover, our data and others’ indicate that clinical benefit is expected in tumors retaining an intact Rb/E2F axis, while resistance-driving alterations (e.g., Rb loss, PIK3CA, ESR1, FGFR1–3, HER2, FAT1 mutations) are likely to limit efficacy. Finally, we highlight cyclin E overexpression as a potential biomarker of resistance to combined CDK4/6i and CDK2i, underscoring the need for biomarker-guided patient stratification. These additions provide a more detailed framework for patient selection in future clinical applications.

References

(1) Finn RS, Crown JP, Lang I, Boer K, Bondarenko IM, Kulyk SO, et al. The cyclin-dependent kinase 4/6 inhibitor palbociclib in combination with letrozole versus letrozole alone as first-line treatment of oestrogen receptor-positive, HER2-negative, advanced breast cancer (PALOMA-1/TRIO-18): a randomised phase 2 study. Lancet Oncol 2015;16:25-35

(2) Finn RS, Martin M, Rugo HS, Jones S, Im S-A, Gelmon K, et al. Palbociclib and Letrozole in Advanced Breast Cancer. New England Journal of Medicine 2016;375:1925-36

(3) Turner NC, Slamon DJ, Ro J, Bondarenko I, Im S-A, Masuda N, et al. Overall Survival with Palbociclib and Fulvestrant in Advanced Breast Cancer. New England Journal of Medicine 2018;379:1926-36

(4) Dickler MN, Tolaney SM, Rugo HS, Cortés J, Diéras V, Patt D, et al. MONARCH 1, A Phase II Study of Abemaciclib, a CDK4 and CDK6 Inhibitor, as a Single Agent, in Patients with Refractory HR(+)/HER2(-) Metastatic Breast Cancer. Clin Cancer Res 2017;23:5218-24

(5) Johnston S, Martin M, Di Leo A, Im S-A, Awada A, Forrester T, et al. MONARCH 3 final PFS: a randomized study of abemaciclib as initial therapy for advanced breast cancer. npj Breast Cancer 2019;5:5

(6) Hortobagyi GN, Stemmer SM, Burris HA, Yap Y-S, Sonke GS, Hart L, et al. Overall Survival with Ribociclib plus Letrozole in Advanced Breast Cancer. New England Journal of Medicine 2022;386:94250

(7) Slamon DJ, Neven P, Chia S, Fasching PA, De Laurentiis M, Im S-A, et al. Overall Survival with Ribociclib plus Fulvestrant in Advanced Breast Cancer. New England Journal of Medicine 2019;382:51424

(8) Im S-A, Lu Y-S, Bardia A, Harbeck N, Colleoni M, Franke F, et al. Overall Survival with Ribociclib plus Endocrine Therapy in Breast Cancer. New England Journal of Medicine 2019;381:307-16

(9) Pandey K, Park N, Park KS, Hur J, Cho YB, Kang M, et al. Combined CDK2 and CDK4/6 Inhibition Overcomes Palbociclib Resistance in Breast Cancer by Enhancing Senescence. Cancers (Basel) 2020;12

(10) Freeman-Cook K, Hoffman RL, Miller N, Almaden J, Chionis J, Zhang Q, et al. Expanding control of the tumor cell cycle with a CDK2/4/6 inhibitor. Cancer Cell 2021;39:1404-21 e11

(11) Dietrich C, Trub A, Ahn A, Taylor M, Ambani K, Chan KT, et al. INX-315, a selective CDK2 inhibitor, induces cell cycle arrest and senescence in solid tumors. Cancer Discov 2023

(12) Al-Qasem AJ, Alves CL, Ehmsen S, Tuttolomondo M, Terp MG, Johansen LE, et al. Co-targeting CDK2 and CDK4/6 overcomes resistance to aromatase and CDK4/6 inhibitors in ER+ breast cancer. NPJ Precis Oncol 2022;6:68

(13) Kudo R, Safonov A, Jones C, Moiso E, Dry JR, Shao H, et al. Long-term breast cancer response to CDK4/6 inhibition defined by TP53-mediated geroconversion. Cancer Cell 2024

(14) Arora M, Moser J, Hoffman TE, Watts LP, Min M, Musteanu M, et al. Rapid adaptation to CDK2 inhibition exposes intrinsic cell-cycle plasticity. Cell 2023;186:2628-43 e21

(15) Kumarasamy V, Wang J, Roti M, Wan Y, Dommer AP, Rosenheck H, et al. Discrete vulnerability to pharmacological CDK2 inhibition is governed by heterogeneity of the cancer cell cycle. Nature Communications 2025;16:1476

(16) Dommer AP, Kumarasamy V, Wang J, O'Connor TN, Roti M, Mahan S, et al. Tumor Suppressors Condition Differential Responses to the Selective CDK2 Inhibitor BLU-222. Cancer Res 2025

(17) Johnson DG, Ohtani K, Nevins JR. Autoregulatory control of E2F1 expression in response to positive and negative regulators of cell cycle progression. Genes & Development 1994;8:1514-25

(18) Chung M, Liu C, Yang HW, Koberlin MS, Cappell SD, Meyer T. Transient Hysteresis in CDK4/6 Activity Underlies Passage of the Restriction Point in G1. Mol Cell 2019;76:562-73 e4

(19) Kim S, Leong A, Kim M, Yang HW. CDK4/6 initiates Rb inactivation and CDK2 activity coordinates cell-cycle commitment and G1/S transition. Sci Rep 2022;12:16810

(20) Yang HW, Chung M, Kudo T, Meyer T, Yang HW, Chung, Mingyu, Kudo T, et al. Competing memories of mitogen and p53 signalling control cell-cycle entry. Nature 2017;549:404-8

(21) Yang C, Li Z, Bhatt T, Dickler M, Giri D, Scaltriti M, et al. Acquired CDK6 amplification promotes breast cancer resistance to CDK4/6 inhibitors and loss of ER signaling and dependence. Oncogene 2017;36:2255-64

(22) Li Q, Jiang B, Guo J, Shao H, Del Priore IS, Chang Q, et al. INK4 Tumor Suppressor Proteins Mediate Resistance to CDK4/6 Kinase Inhibitors. Cancer Discov 2022;12:356-71

(23) Ji W, Zhang W, Wang X, Shi Y, Yang F, Xie H, et al. c-myc regulates the sensitivity of breast cancer cells to palbociclib via c-myc/miR-29b-3p/CDK6 axis. Cell Death & Disease 2020;11:760

(24) Wu X, Yang X, Xiong Y, Li R, Ito T, Ahmed TA, et al. Distinct CDK6 complexes determine tumor cell response to CDK4/6 inhibitors and degraders. Nature Cancer 2021;2:429-43

(25) Kim S, Son E, Park HR, Kim M, Yang HW. Dual targeting CDK4/6 and CDK7 augments tumor response and anti-tumor immunity in breast cancer models. J Clin Invest 2025

(26) Ravani LV, Calomeni P, Vilbert M, Madeira T, Wang M, Deng D, et al. Efficacy of Subsequent Treatments After Disease Progression on CDK4/6 Inhibitors in Patients With Hormone Receptor-Positive Advanced Breast Cancer. JCO Oncol Pract 2025;21:832-42

(27) Martin JM, Handorf EA, Montero AJ, Goldstein LJ. Systemic Therapies Following Progression on Firstline CDK4/6-inhibitor Treatment: Analysis of Real-world Data. Oncologist 2022;27:441-6

(28) Kalinsky K, Bianchini G, Hamilton E, Graff SL, Park KH, Jeselsohn R, et al. Abemaciclib Plus Fulvestrant in Advanced Breast Cancer After Progression on CDK4/6 Inhibition: Results From the Phase III postMONARCH Trial. J Clin Oncol 2025;43:1101-12

Version 2

Abstract

CDK4/6 inhibitors (CDK4/6i) with endocrine therapy are standard for hormone receptor-positive (HR+) metastatic breast cancer. However, most patients eventually develop resistance and discontinue treatment, and there is currently no consensus on effective second-line strategies. Here, we demonstrate that maintaining CDK4/6i therapy, either alone or in combination with CDK2 inhibitors (CDK2i), slows the growth of drug-resistant HR+ models by prolonging G1 progression. Mechanistically, sustained CDK4/6 blockade in drug-resistant cells reduces E2F transcription and delays G1/S via a non-canonical, post-translational regulation of retinoblastoma protein (Rb). Durable suppression of both CDK2 activity and growth of drug-resistant cells requires co-administration of CDK2i with CDK4/6i. Moreover, cyclin E overexpression drives resistance to the combination of CDK4/6i and CDK2i. These findings elucidate how continued CDK4/6 blockade constrains resistant tumors and support clinical strategies that maintain CDK4/6i while selectively incorporating CDK2i to overcome resistance.

Introduction

Metastatic breast cancer remains a leading cause of cancer-related mortality in women globally (1,2). A key dysregulation in breast cancer involves the overactivation of cyclin-dependent kinases 4 and 6 (CDK4/6) (3–6). Active CDK4/6 phosphorylates the retinoblastoma protein (Rb), a crucial regulator that prevents cell-cycle initiation by sequestering E2F transcription factors (7). Rb phosphorylation results in the release of E2F, thereby promoting CDK2 activation and cell proliferation (8,9). Understanding this mechanism has driven significant advancements in therapeutic strategies, particularly for hormone receptor-positive (HR+)/human epidermal growth factor receptor 2-negative (HER2-) breast cancer, which constitutes approximately 70% of breast cancer cases (10). The current standard first-line treatment for HR+/HER2- metastatic breast cancer is a combination of CDK4/6 inhibitors (CDK4/6i) and endocrine therapy (ET) (4,5,11). Although this strategy has significantly improved patient outcomes, resistance remains a major challenge: approximately 30% of patients develop resistance within two years, and the majority ultimately relapse (12,13). Upon disease progression, CDK4/6i therapy is typically discontinued, often resulting in aggressive tumor regrowth and a lack of effective second-line treatment options.

While continuing ET after disease progression confers clinical benefit (14–18), the value of maintaining CDK4/6i treatment beyond progression remains elusive. Multiple ongoing and completed trials have evaluated continuing the FDA-approved CDK4/6 inhibitors, palbociclib, abemaciclib, and ribociclib, after progression (19–23). MAINTAIN and postMONARCH trials reported significantly improved progression-free survival (PFS) with continued CDK4/6i plus ET (21,23). Retrospective analyses likewise suggest that second-line CDK4/6i plus ET can outperform chemotherapy or ET alone in several settings (24,25). Safety appears comparable or better than first-line trials, with no new toxicities and grade 3–4 events largely confined to expected hematologic adverse effects. In contrast, the PACE and PALMIRA trials did not demonstrate a PFS benefit for continuing CDK4/6i plus ET after progression (19,20). Overall, the evidence for CDK4/6i continuation beyond progression is encouraging but mixed, underscoring the need for mechanistic studies of CDK4/6i-resistant proliferation to guide post-progression treatment strategies for HR+/HER2- breast cancer.

Like CDK4/6, CDK2 also phosphorylates Rb before the G1/S transition (8,26). The relevance of CDK2 activation in CDK4/6i-resistant tumors highlights the potential of targeting CDK2 to overcome CDK4/6i resistance (27–34). Several clinical trials are currently evaluating CDK2 inhibitors (CDK2i) in patients who have progressed on CDK4/6i-based therapy, either as monotherapy or in combination with CDK4/6i, with or without ET (NCT05252416, NCT05735080). Determining the optimal strategy for incorporating CDK2i in treating HR+/HER2-breast cancer that has developed drug resistance remains a critical question.

This study shows that, even in drug-resistant cells that remain proliferative, continued CDK4/6 inhibition results in ineffective Rb inactivation, thereby slowing E2F activation kinetics and prolonging the G1 phase. These data suggest that overall survival, rather than PFS, may serve as a more appropriate clinical trial endpoint. Moreover, our work highlights the therapeutic potential of combining CDK2i with CDK4/6i as an effective second-line strategy. Finally, we identify cyclin E overexpression as a key driver of resistance to this combination, offering mechanistic guidance for overcoming therapy resistance.

Results

Maintaining CDK4/6i treatment attenuates the growth of drug-resistant cells by extending G1-phase progression

To evaluate the impact of continued CDK4/6i treatment in drug-resistant settings, we employed HR+/HER2- breast cancer cell lines (MCF-7, T47D, and CAMA-1) and a triple-negative breast cancer cell line (MDA-MB-231), all of which have an intact Rb/E2F pathway. These cells were chronically exposed to palbociclib for over a month to induce drug resistance (Figure 1A). We confirmed increased half-maximal inhibitory concentrations (IC50) for palbociclib in drug-resistant cells compared to parental cells (Figure S1A). We maintained or withdrew CDK4/6i treatment in drug-resistant cells and monitored their cumulative proliferation alongside that of parental cells. Drug-resistant cells maintained on CDK4/6i exhibited significantly slower growth than those under other conditions (Figure 1B). In contrast, resistant cells withdrawn from treatment grew at rates comparable to parental cells. Thus, despite the emergence of resistance, continued CDK4/6i treatment markedly suppresses the growth rate of drug-resistant cells.

Continuous CDK4/6i treatment suppresses the growth of drug-resistant cells.

(A) Schematic illustrating the establishment of CDK4/6i-resistant cells, either maintained in drug or withdrawn from treatment, for use in subsequent experiments. (B) Growth curves of parental and drug-resistant cells. Palbociclib (1 µM) was either withdrawn or maintained in drug-resistant cells. Data represent mean ± SD (n = 3 biological replicates). Statistical significance was determined by two-way ANOVA with Tukey’s post-hoc analysis (**** P < 0.0001). (C) Schematic of live-cell reporters for CDK4/6 and CDK2 activities (top) and cell-cycle phase (bottom). (D, E) Representative single-cell traces showing CDK4/6 and CDK2 activities together with Cdt1-degron intensity in parental (D) and drug-resistant (E) MCF-7 cells. Resistant cells were maintained in palbociclib (1 µM). (F, G) Heatmaps of single-cell traces showing CDK4/6 (left) and CDK2 (middle) activities, and Cdt1-degron intensity (right) in drug-resistant cells without (F) or with (G) continuous palbociclib (1 µM). Proliferating cells were identified by sustained CDK2 activity (>1 for >2 hr during 30–48 hr). (H, I) Single-cell traces of CDK2 activity (left) and Cdt1-degron intensity (right), aligned to the end of mitosis (anaphase) in drug-resistant cells without (H) or with (I) continuous palbociclib (1 µM). Based on CDK2 activity (black line), cells were classified as proliferating (red) or quiescent (blue).

To investigate how continued CDK4/6i treatment impedes the growth of drug-resistant cells, we monitored cell-cycle progression at the single-cell level. We established MCF-7 and MDA-MB-231 cells expressing live-cell sensors for CDK4/6 activity (35), CDK2 activity (36), and cell-cycle phase (Cdt1 degron) (37), as well as a nucleus marker (histone 2B) for individual cell tracking. The CDK sensors dynamically shuttle between the nucleus and cytoplasm depending on phosphorylation by their respective kinases (Figure 1C top). The Cdt1 degron is degraded during the S phase, allowing for the visualization of the G1/S and S/G2 transitions (Figure 1C, bottom). In parental cells, CDK4/6 was continuously activated after mitosis, followed by a gradual activation of CDK2, leading to S-phase entry (Figure 1D). In contrast, drug-resistant cells entered S phase through CDK2 activation with minimal CDK4/6 activity (Figure 1E). When evaluating thousands of cells, about 98% of parental cells exhibited sustained CDK4/6 and CDK2 activation to enter the cell cycle (Figure S1C and S1D). While drug withdrawal reactivated CDK4/6, over 90% of drug-resistant cells continued to proliferate regardless of the presence of CDK4/6i (Figure 1F, 1G, S1E, and S1F). To further assess cell-cycle dynamics, we aligned cells at cytokinesis and classified them as proliferating or quiescent based on CDK2 activity. In the absence of CDK4/6i, drug-resistant cells displayed CDK2/46 and CDK2 activity dynamics and S-phase entry kinetics comparable to those of parental cells (Figure 1H and S2A–S2D). By contrast, continued CDK4/6i treatment in resistant cells resulted in persistently low CDK4/6 activity, delayed CDK2 activation kinetics, and greater heterogeneity in the timing of S-phase entry, as indicated by Cdt1 degradation (Figure 1I, S2E, and S2F). These results demonstrate that despite active proliferation, continuous CDK4/6i treatment markedly decelerates and destabilizes cell-cycle progression in resistant cells, thereby limiting their growth rates.

CDK6 overexpression has emerged as a key driver of CDK4/6i resistance (38–41). In line with these findings, we observed elevated CDK6 levels in CDK4/6i-resistant cells (Figure S3A). To investigate the role of CDK6 in modulating CDK4/6 activity under resistance, we established CDK6 knockout (KO) MCF-7 cells using CRISPR-Cas9 and induced resistance to CDK4/6i (Figure S3B). Despite CDK6 depletion, CDK4/6 activity remained comparable between wild-type and CDK6-KO cells under continuous CDK4/6i treatment (Figure S3C–S3F). These results suggest a potential non-canonical role for CDK6 in mediating resistance to CDK4/6i therapy.

To assess the impact of continuous CDK4/6i treatment on cell-cycle progression, we measured the intermitotic time and the duration of each cell-cycle phase in MCF-7 and CAMA-1 cells (Figure 2A). The intermitotic time was comparable between parental and drug-resistant cells without ongoing CDK4/6i treatment (Figure 2B). However, in drug-resistant cells that were continuously treated with CDK4/6i, the intermitotic time was extended by 30–50% compared to other conditions. Moreover, continuous CDK4/6i treatment prolonged the G1 phase by approximately 200–300% without affecting the duration of the S and G2/M phases (Figure 2C–2E). Our data indicate that maintaining CDK4/6i treatment in drug-resistant cells significantly extends the G1 phase, thereby decelerating the overall growth rate.

CDK4/6i maintenance extends G1-phase progression.

(A) Representative single-cell traces showing CDK4/6 and CDK2 activities (top) and Cdt1-degron intensity (bottom) in parental MCF-7 cells. (B–E) Violin plots showing intermitotic time (n > 600 cells/condition) (B), G1-phase duration (n > 2,000 cells/condition) (C), S-phase duration (n > 400 cells/condition) (D), and G2/M-phase duration (n > 250 cells/condition) (E) in MCF-7 (left) and CAMA-1 (right) cells. Solid and dashed yellow lines indicate mean and median, respectively.

Maintaining CDK4/6i treatment triggers ineffective Rb inactivation

We sought to elucidate the molecular mechanisms underlying the extended G1 phase and slow CDK2 activation kinetics in drug-resistant cells under continuous CDK4/6i treatment. Our recent studies demonstrated that CDK4/6 inhibition initially halts proliferation via Rb activation but ultimately reduces Rb protein due to decreased stability (42–44). Consistent with this, we found that drug-resistant cells maintained on CDK4/6i exhibited reduced levels of both total and phosphorylated Rb compared to parental cells (Figure 3A). Importantly, CDK4/6i withdrawal restored both total and phosphorylated Rb levels, confirming that the effect was reversible. However, Rb loss in resistant cells was incomplete, as demonstrated by comparisons with Rb-KO cells (Figure S4A). To directly test the role of Rb, we used Rb-KO cells resistant to CDK4/6i. Remarkably, Rb KO rescued the prolonged G1 duration and restored rapid CDK2 activation kinetics without affecting CDK4/6 activity, even under continuous CDK4/6 inhibition (Figures S4B and S4C). These findings led us to hypothesize that maintaining CDK4/6i induces a non-canonical pathway of Rb inactivation via post-translational degradation. This passive inactivation may result in inefficient E2F activation and delayed CDK2 engagement, thereby slowing G1-phase progression.

CDK4/6i maintenance induces an ineffective Rb inactivation pathway.

(A) Immunoblot of phosphorylated Rb (S807/811; p-Rb), total Rb (t-Rb), and β-actin protein levels in MCF-7 cells. Drug-resistant cells were harvested two weeks after drug withdrawal. (B) Singe-cell traces of CDK2 activity aligned to mitosis in proliferating (red) and quiescent (blue) MCF-7 cells. Circles indicate the time of fixation and staining (n = 200 cells). (C) Scatter plot of EdU intensity versus time since mitosis in parental cells. The red dotted line marks the threshold for S phase (n = 2,000 cells). (D, E) Percentage of S-phase cells as a function of time since mitosis in proliferating (D) and quiescent (E) cells. Data represent mean ± SD (n = 2 biological replicates). (F) Representative immunostaining of Hoechst, p-Rb, t-Rb, and E2F1 mRNA FISH (top). Processed images showing nucleus segmentation, p-Rb classification, and mRNA puncta detection (bottom). (G) Histogram of cells with p-Rb normalized to t-Rb. (H, I) Percentage of cells with p-Rb (H) and E2F1 mRNA levels (I) as a function of time since mitosis. Data represent mean ± SD (H, n = 3 biological replicates) or mean ± 95% confidence intervals (I, n > 2,500 cells/condition). (J, K) E2F1 mRNA (J) and EdU (K) levels as a function of CDK2 activity. Data represent mean ± 95% confidence intervals (n > 2,500 cells/condition). (L, M) Averaged traces of CDK4/6 (left) and CDK2 (middle) activities and Cdt1-degron intensity (right) (L) and percentage of S-phase cells (M). MCF-7 cells were released after synchronizing with nocodazole (250 nM, 14 hr).

To test this hypothesis, we tracked CDK2 activity in individual cells and aligned proliferative events with fixed-cell measurements from the end of mitosis (Figure 3B and S5A). Using 5-ethynyl-2’-deoxyuridine (EdU) incorporation, we quantified the kinetics of S-phase entry. Parental and drug-resistant cells without CDK4/6i transitioned sharply into S phase approximately 6 hr after mitosis and completed DNA replication within 12 hr (Figures 3C, 3D, and S5B). In contrast, drug-resistant cells under sustained CDK4/6 inhibition exhibited markedly delayed and heterogeneous G1/S transitions (Figures 3D and S5C). Quiescent cells, as expected, showed no EdU incorporation (Figure 3E). We next assessed the dynamics of Rb phosphorylation and E2F activity using immunostaining and mRNA FISH (Figure 3F). Rb hyperphosphorylation was classified by phosphorylation at Ser807/811, a well-characterized marker (45) (Figure 3G). E2F1 mRNA levels, which are autoregulated (46), serve as a proxy for E2F transcriptional output (8,45,47). In the absence of CDK4/6i, both parental and drug-resistant cells robustly induced Rb phosphorylation before S-phase entry (Figure 3H). Conversely, under continuous CDK4/6 inhibition, resistant cells initiated CDK2 activation and G1 progression without Rb hyperphosphorylation, resulting in slower and weaker E2F1 mRNA induction (Figures 3H and 3I). When comparing CDK2 activity with E2F1 mRNA levels or EdU incorporation, resistant cells under CDK4/6i maintenance required higher CDK2 activity to reach equivalent E2F1 mRNA levels or S-phase entry (Figure 3J, 3K, S5D, and S5E). The reduced E2F1 mRNA levels during S phase likely result from suppression by atypical E2Fs, E2F7 and E2F8 (48). Finally, mitotic synchronization with nocodazole confirmed that CDK4/6i maintenance prolongs G1 and delays S-phase entry in resistant cells (Figure 3L and 3M). Together, these findings indicate that continuous CDK4/6 inhibition in drug-resistant cells triggers inefficient Rb inactivation, attenuated E2F induction, and delayed CDK2-driven G1/S transition, leading to prolonged and heterogeneous G1-phase progression.

Maintenance of CDK4/6i suppresses the growth of drug-resistant tumors

We investigated the therapeutic benefits of maintaining CDK4/6i treatment after the onset of drug resistance in vivo. To this end, we established an MCF-7 xenograft model by orthotopically inoculating MCF-7 cells into the mammary fat pad of immunodeficient mice. Palbociclib was initially administered when tumors reached 100 mm3 (Figure 4A). Following an initial period of tumor stasis/regression of variable duration, tumors resumed growth, indicating the development of acquired resistance (Figure 4B). Upon regrowth to 145–155 mm3, mice were randomized to one of four secondary regimens: treatment discontinuation, continued palbociclib, or a switch to ribociclib or abemaciclib. To mirror clinical practice, ribociclib was administered at a fourfold higher dose than the other CDK4/6i. Consistent with our in vitro findings, continuation of any CDK4/6i significantly attenuated overall tumor growth relative to treatment discontinuation (Figure 4C–E). Additionally, we observed that switching to ribociclib or abemaciclib resulted in significantly greater tumor control compared to continuing palbociclib. These observations align with clinical studies reporting agent-specific differences in second-line PFS (19–21,23). Notably, analysis of tumor tissues after secondary treatment showed restoration of total Rb expression only in the drug-discontinued cohort (Figure 4F). Together, these data support maintaining CDK4/6 inhibition to restrain the overall growth of drug-resistant tumors.

CDK4/6i maintenance suppresses the growth of drug-resistant tumors.

(A) Schematic representation of experimental design. Once tumors reached a volume of 100 mm3, mice were treated with palbociclib. Following the development of resistance, mice were randomly assigned to one of four treatment groups: treatment discontinuation, palbociclib maintenance, switch to ribociclib, or switch to abemaciclib for 32 days. (B) Tumor growth curves showing the establishment of resistance to palbociclib. Horizontal red dotted lines (145−155 mm3) indicate the point at which mice were assigned to second CDK4/6i treatment (n = 20 mice). (C, D) Averaged (C) and individual (D) tumor growth traces following second CDK4/6i treatment. Data represent mean ± SEM (n = 5 mice/condition). Statistical significance was determined using two-way ANOVA with Tukey’s post-hoc analysis (**** P < 0.0001). (E, F) Box plots showing tumor mass (E) and the percentage of Rb-positive cells (F) after second CDK4/6i treatment (n = 5 mice/condition). Statistical significance was determined using one-way ANOVA with Tukey’s post-hoc analysis (** P < 0.05, *** P < 0.01, **** P < 0.0001).

Addition of ET augments the efficacy of CDK4/6i maintenance

Our recent studies have highlighted the pivotal role of c-Myc in amplifying E2F activity and fostering CDK4/6i resistance following alternative Rb inactivation (42,43). Using a doxycycline-inducible system, we found that induction of c-Myc significantly facilitated the growth of drug-resistant cells under ongoing CDK4/6i treatment (Figure S6A). Furthermore, despite similar low CDK4/6 activity, c-Myc induction facilitated cell-cycle progression in CDK4/6i-resistant cells (Figure S6B–D). These data demonstrated that c-Myc promotes CDK4/6i resistance by facilitating cell-cycle progression in drug-resistant cells.

Given the estrogen responsiveness of c-Myc (49), we hypothesized that the continued addition of ET could further suppress the growth of CDK4/6i-resistant cells by downregulating c-Myc expression. We employed the estrogen receptor antagonist fulvestrant to evaluate the therapeutic benefit of maintaining the combination of CDK4/6i and ET. While MCF-7 cells exhibited primary resistance to fulvestrant, CAMA-1 and T47D cells displayed sensitivity to this drug (Figure 5A). We used these MCF-7 and CAMA-1 cells to examine whether primary ET resistance contributes to the benefit of maintaining CDK4/6i and ET in drug-resistant cells. To establish drug resistance, CDK4/6i-resistant cell lines were chronically treated with a combination of palbociclib and fulvestrant for over two months. Subsequently, these resistant cell lines were subjected to continuous treatment with either palbociclib or fulvestrant alone or in combination, and their cumulative proliferation rate was monitored. We observed a significantly slower growth rate in drug-resistant cells continuously exposed to the combination therapy compared to other treatment conditions, not only in CAMA-1 but also in MCF-7 cells (Figure 5B, S7A, and S7B). Continued treatment with fulvestrant, either alone or combined with palbociclib, reduced c-Myc levels compared to palbociclib maintenance alone (Figure 5C). Our data indicate distinct resistance mechanisms between CDK4/6i and ET. Therefore, regardless of primary resistance to ET, the continued addition of ET to CDK4/6i treatment is beneficial in suppressing the growth of drug-resistant cells by inhibiting c-Myc.

Maintaining CDK4/6i and ET synergistically suppresses the growth of drug-resistant cells.

(A) IC50 values of fulvestrant. Data represent mean ± SD (n = 3 biological replicates). Statistical significance was determined by one-way ANOVA with Tukey’s post-hoc analysis (*** P < 0.001). Growth curves of MCF-7 (left) and CAMA-1 (right) cells resistant to palbociclib and fulvestrant. Cells were maintained with fulvestrant (500 nM), palbociclib (1 µM), or the combination. Data represent mean ± SD (n = 3 biological replicates). Statistical significance was determined by two-way ANOVA with Tukey’s post-hoc analysis (**** P < 0.0001). (C) Violin plots of c-Myc levels measured by immunofluorescence in resistant cells resistant treated with the indicated drugs for one week. Solid and dashed yellow lines represent mean and median, respectively (n > 2,000 cells/condition). (D, E) Heatmaps showing single-cell traces for CDK4/6 (left) and CDK2 (middle) activities and Cdt1-degron intensity (right) in drug-resistant cells maintained with palbociclib (1 µM) and fulvestrant (500 nM) (D) or fulvestrant alone (E). Proliferating cells were identified based on CDK2 activity (>1 for more than 2 hr between 30 and 48 hr). (F) Averaged traces of CDK4/6 (left) and CDK2 (right) activities aligned by mitosis in drug-resistant cells treated with the indicated drugs. Data represent mean ± 95% confidence intervals (n > 1,000 cells/condition). (G) Single-cell traces of Cdt1-degron intensity aligned by mitosis in drug-resistant cells treated with palbociclib (1 µM) (left) or palbociclib + fulvestrant (500 nM) (right). (H, I) Violin plots showing intermitotic time (n > 70 cells/condition) (H), and G1-phase duration (n > 630 cells/condition) (I) in MCF-7 (left) and CAMA-1 (right) cells. Solid and dashed yellow lines indicate mean and median, respectively.

To elucidate the mechanisms underlying the attenuation of growth resulting from continued treatment with the drug combination, we assessed CDK4/6 and CDK2 activities along with cell-cycle progression (Cdt1 degron) in drug-resistant MCF-7 cells. Cells were continuously treated with palbociclib or fulvestrant alone or in combination. By classifying cells into proliferation and quiescence based on CDK2 activation, we found over 80% of cells were proliferating in all conditions (Figure 5D, 5E, and S7C). Furthermore, we observed reactivation of CDK4/6 in drug-resistant cells continuously treated with fulvestrant alone. However, when we monitored cell-cycle dynamics by aligning cells to mitosis, the maintenance of the drug combination caused slower CDK2 activation kinetics and greater heterogeneity in S-phase entry, as indicated by Cdt1 degradation, compared to single-drug treatment conditions (Figure 5F, 5G, S7D, and S7E). By analyzing each cell-cycle duration, we observed a significant increase in intermitotic time and G1-phase duration, while S- or G2-phase lengths remained relatively unchanged (Figure 5H, 5I, S7F, and S7G). This emphasizes the role of the drug combination in altering cell-cycle kinetics in the G1 phase. Our findings showed the therapeutic benefits of maintaining a combination of CDK4/6i and ET in attenuating the growth of drug-resistant cells.

The benefit of combining CDK2i with CDK4/6i as a second-line therapy

We next evaluated the therapeutic potential of adding the CDK2i INX-315 as a second-line strategy. Palbociclib/fulvestrant-resistant MCF-7 and CAMA-1 cells were treated under five regimens: treatment discontinuation, fulvestrant alone, fulvestrant + palbociclib, fulvestrant + INX-315, or fulvestrant + palbociclib + INX-315. Continuous palbociclib + fulvestrant treatment significantly reduced resistant-cell growth compared to discontinuation, fulvestrant alone, or fulvestrant + INX-315 (Figure 6A–6C). Notably, the triple combination provided the most robust suppression of growth. To dissect cell-cycle dynamics, we assessed CDK activities and Cdt1 degron levels one week after treatment. Consistent with earlier results, palbociclib withdrawal triggered CDK4/6 reactivation (Figure S8A–S8D). Furthermore, we found CDK2 reactivation even in the presence of INX-315 (Figure 6D and S8D). While CDK2 reactivation was also observed, the triple combination produced the longest intermitotic intervals and G1-phase duration without affecting the S and G2 phases (Figure 6E and S8E). Analysis of cells synchronized by mitosis further confirmed that the triple-drug regimen most strongly delayed CDK2 activation kinetics (Figure 6F). Moreover, dose titration studies revealed a synergistic interaction between CDK4/6i and CDK2i (Figure 6G and S8F). To explore transcriptomic consequences, we performed RNA sequencing 20 days post-treatment. Continued palbociclib + fulvestrant suppressed E2F target genes more effectively than INX-315, and the triple combination achieved the most potent inhibition of both E2F and Myc target gene programs (Figure 6H and S8G). Together, these findings demonstrate that, despite partial CDK2 reactivation, adding CDK2i to ongoing CDK4/6i + ET further attenuates E2F activity, delays CDK2 engagement, and suppresses resistant-cell growth. These results support a therapeutic rationale for maintaining CDK4/6 blockade while strategically incorporating CDK2 inhibition after progression.

Therapeutic benefit of combining CDK2i with CDK4/6i and ET.

(A, B) Growth curves of MCF-7 (left) and CAMA-1 (right) cells resistant to palbociclib and fulvestrant under various treatment conditions: drug discontinuation, fulvestrant (500 nM) alone, or in combination with palbociclib (1 µM) and/or INX-315 (100 nM). Data represent mean ± SD (n = 3 biological replicates). P values were calculated with two-way ANOVA with Tukey’s post-hoc analysis (**** P < 0.0001). (C) Cell numbers 20 days after drug treatment. Data represent mean ± SD (n = 3 biological replicates). P values were calculated using an unpaired t-test (* P < 0.05, ** P < 0.01, *** P < 0.001, **** P < 0.0001). (D) Heatmaps of single-cell traces for CDK4/6 (left) and CDK2 (middle) activities, and Cdt1-degron intensity (right) in palbociclib/fulvestrant resistant MCF-7 cells treated with the triple combination of palbociclib (1 µM), fulvestrant (500 nM), and INX-315 (100 nM) for one week before imaging. Proliferating cells were identified based on CDK2 activity (>1 for more than 2 hr between 30 and 48 hr). (E) Violin plots showing intermitotic time (n > 200 cells/condition) (left) and G1-phase duration (n > 900 cells/condition) (right). Solid and dashed yellow lines indicate mean and median, respectively. (F) Averaged traces of CDK4/6 (left) and CDK2 (right) activities aligned by mitosis in drug-resistant cells treated with the indicated drugs. Data represent mean ± 95% confidence intervals (n > 1,000 cells/condition). (G) Loewe synergy score calculated from percent-inhibition data generated by dual titration of palbociclib (0–10µM) and INX-315 (0–30µM) in MCF-7 cells following 48 hr treatment (n = 3 biological replicates). (H) Heatmaps comparing gene expression profiles for hallmark E2F (left) and MYC (right) targets in drug-resistant MCF-7 cells treated with the indicated drugs for 20 days. Samples were collected as biological duplicates.

Role of cyclin E/A in CDK2 reactivation under combined CDK4/6 and CDK2 inhibition

A recent study identifies the cyclin A-CDK complex as a key driver of rapid adaptation to CDK2 inhibition (32). To investigate the mechanism of CDK2 reactivation during CDK2i treatment, we explored the role of the CDK2 activators, cyclins E and A. Using a doxycycline-inducible system, we induced the overexpression of cyclin E1 or A2 in drug-naïve MCF-7 cells (Figure S9A). While treatment with INX-315 (100 nM–1 µM) initially inhibited CDK2 activity, CDK2 reactivation occurred within 1–2 hr after drug treatment, regardless of cyclin E1 or A2 overexpression (Figure 7A and S9B–S9D). The combination of CDK2i and CDK4/6i blocked this rapid CDK2 reactivation, but the overexpression of cyclin E1 or A2 hindered complete suppression (Figure 7B and S9E–S9G). These findings suggest that CDK4/6 inhibition is necessary for effectively suppressing CDK2 activity by CDK2i, and that cyclin E1 and A2 overexpression may contribute to CDK2 reactivation.

Overexpression of cyclin E and A facilitates resistance to the combination of CDK2i and CDK4/6i.

(A, B) Averaged traces of CDK4/6 (left) and CDK2 (right) activities in MCF-7 cells with or without cyclin E1 or A2 overexpression. Cells were treated with doxycycline (500 nM) 6 hr before the addition of INX-315 (1 µM) alone (A) or in combination with palbociclib (1 µM) (B). Data represent mean ± 95% confidence intervals (n > 900 cells/condition). (C, D) Heatmaps of single-cell traces for CDK4/6 (left) and CDK2 (middle) activities, and Geminin-degron intensity (right) in drug-naïve MCF-7 cells without (C) or with (D) cyclin E1 overexpression. Cells were treated with palbociclib (1 µM), INX-315 (100 nM), and doxycycline (500 nM). Persister cells were identified based on CDK2 activity (>1 for more than 2 hr between 30 and 48 hr). (E) Averaged traces of CDK4/6 (left) and CDK2 (middle) activities and Geminin degron intensity (right) in persister cells tolerant to combination palbociclib (1 µM), and INX-315 (100 nM) without or with cyclin E1 or A2 overexpression. Data represent mean ± 95% confidence interval. n > 2500 cells/condition. (F) Growth curves of palbociclib-resistant MCF-7 cells without or with cyclin E1 or A2 overexpression. Cells were treated with palbociclib (1 µM), INX-315 (100 nM), and doxycycline (500 nM). Data represent mean ± SD (n = 3 biological replicates). Statistical significance was determined using two-way ANOVA with Tukey’s post-hoc analysis (**** P < 0.0001). (G) Summary schematic illustrating the mechanisms underlying the benefits of continued CDK4/6i and ET therapies and the introduction of CDK2i in drug-resistant cells.

We next examined the impact of cyclin E and A overexpression in the development of persister cells, which are implicated in residual tumor growth and eventual resistance (42,50). We used drug-naïve MCF-7 cells expressing live-cell sensors for CDK2, CDK4/6, and anaphase-promoting complex/cyclosome (APC/C) (Geminin degron) activities (51). APC/C is a multi-subunit E3 ubiquitin ligase typically inactivated near the G1/S transition, leading to Geminin-degron accumulation. Treatment with CDK4/6i and CDK2i induced near-complete cell-cycle arrest within 24 hr (Figure 7C). However, approximately 10% of cells developed a persister phenotype through CDK2 reactivation. Overexpression of cyclin E1 increased the percentage of persister cells, while cyclin A2 had no obvious impact (Figure 7D and S10). Further kinetic analysis revealed that cyclin E1 overexpression accelerated CDK2 reactivation and APC/C inactivation in persister cells (Figure 7E). These findings indicate that cyclin E1, but not cyclin A2, facilitates CDK2 reactivation and persister development, promoting resistance to the CDK4/6i and CDK2i combination.

Since cyclin A is a substrate of APC/C (52), the lack of effect from cyclin A2 induction could be due to its degradation by APC/C. Thus, this degradation may reduce cyclin A to insufficient levels, preventing its impact on persister cell development. To further evaluate the impact of cyclin E and A overexpression on resistance development, we exposed palbociclib-resistant MCF-7 cells to the combination of palbociclib and INX-315. Both cyclin E1 and A2 overexpression significantly accelerated the development of resistance to the drug combination, with cyclin E1 having a more significant effect than cyclin A2 (Figure 7F). These results indicate that although CDK2 activity is targeted by CDK2i, the overexpression of cyclin E and A promotes resistance to the combination of CDK4/6i and CDK2i. Due to cyclin A degradation by APC/C in quiescent cells, cyclin E may play a more critical role in driving resistance to this drug combination.

Discussion

The combination of CDK4/6i and ET has reshaped treatment for HR+/HER2- breast cancer (53–60). However, resistance commonly emerges, and no consensus second-line standard is established. Our data show that continued CDK4/6i treatment in drug-resistant cells engages a non-canonical, proteolysis-driven route of Rb inactivation, yielding attenuated E2F output and a pronounced delay in G1 progression (Figure 7G). Concurrent ET further deepens this blockade by suppressing c-Myc-mediated E2F amplification, thereby prolonging G1 and slowing population growth. Importantly, CDK2 inhibition alone was insufficient to control resistant cells. Robust suppression of both CDK2 activity and resistant-cell growth required CDK2i in combination with CDK4/6i, consistent with prior reports supporting dual CDK targeting (27–34). Moreover, cyclin E blunted the efficacy of the CDK4/6i + CDK2i combination by reactivating CDK2. Together, these findings provide a mechanistic rationale for maintaining CDK4/6i beyond progression and support testing the combination of CDK4/6i and CDK2i, as evidenced by concordant in vitro and in vivo results.

Our data indicate that maintaining both CDK4/6i and ET synergistically decelerates cell-cycle progression in drug-resistant cells by further delaying CDK2 activation kinetics and the G1/S transition without affecting the S and G2 phases. This dual effect stems from CDK4/6i causing suboptimal Rb inactivation while ET suppresses the global transcription amplifier c-Myc, collectively leading to diminished E2F transcriptional activity. As a result, this reduced E2F activity lowers the expression of critical cell-cycle genes, such as cyclin E and A, extending the time needed for CDK2 activation. Given that CDK2 plays an essential role in initiating and advancing DNA replication (61,62), its delayed activation significantly prolongs the G1/S transition. Moreover, CDK2 activation also contributes to Rb phosphorylation and inactivation. High CDK2 activity is required to phosphorylate Rb, and CDK2-mediated Rb phosphorylation is tightly coupled with DNA replication timing (8,45). Thus, upon Rb phosphorylation by CDK2 at the G1/S transition, drug-resistant cells may effectively proceed through the cell cycle even under continued CDK4/6i treatment.

Clinical trials evaluating the efficacy of sustained CDK4/6i therapy predominantly use PFS as the primary endpoint (19–23). However, our findings suggest that drug-resistant tumors continue to proliferate despite CDK4/6i maintenance. Consequently, maintaining CDK4/6i appears to slow tumor growth rather than completely arrest it. This underscores the need for clinical trials to consider overall survival and tumor progression rates as more appropriate endpoints for assessing the true benefits of sustained CDK4/6i therapy. Furthermore, the distinct polypharmacology profiles among CDK4/6i (63), with ribociclib being the most specific and abemaciclib the least, may explain the varying therapeutic outcomes observed among these inhibitors (23,64).

Maintaining CDK4/6i treatment beyond progression may be particularly beneficial for about 70% of patients who do not acquire new genetic mutations (65). However, it is important to recognize that resistance to CDK4/6i often arises from mutations in genes associated with mitogenic or hormone-signaling pathways (65–69). These include mutations in PIK3CA, ESR1, FGFR1–3, and HER2, which have been linked to increased c-Myc expression (49,70,71). Additionally, previous studies have identified FAT1 mutations as a driver of CDK4/6i resistance (39,72). These resistance mutations may reduce the efficacy of maintaining CDK4/6i and ET therapy. Moreover, about 4.7% of HR+/HER2- breast cancer patients exhibit Rb mutations (65,67), making CDK4/6i treatment unlikely to be effective, thus making its continuation inadvisable in these cases.

In conclusion, our study provides mechanistic rationale for maintaining CDK4/6i together with ET after disease progression in HR+/HER2- breast cancers that retain an intact Rb/E2F pathway. The combination of CDK4/6i and CDK2i can further provide durable growth suppression, consistent with prior studies (27–34). However, it is essential to acknowledge that CDK2/4/6 inhibition may promote whole-genome duplication (73), potentially fueling more aggressive tumor evolution. Finally, we identify cyclin E overexpression as a key driver of resistance to dual CDK4/6i and CDK2i therapy, providing a basis for biomarker-guided patient selection and the development of strategies to overcome therapeutic resistance.

Materials and Methods

Cell culture

MCF-7 (ATCC, HTB-22), CAMA-1 (ATCC, HTB-21), and MDA-MB-231 (ATCC, CRM-HTB-26) cells were cultured in Dulbecco’s Modified Eagle Medium (DMEM; Genesee Scientific, 25-500) supplemented with 10% fetal bovine serum (FBS; Gibco, A3160601). T47D (ATCC, HTB-133) cells were maintained in RPMI-1640 (Genesee Scientific, 25-206) with 10% FBS. All cell lines were grown at 37 °C in a humidified incubator with 5% CO₂ and were routinely tested and confirmed negative for mycoplasma contamination.

Drugs and Chemicals

For in vitro experiments, palbociclib (Selleck Chemicals, S1116), INX-315 (MedChemExpress, HY-162001), and fulvestrant (Selleck Chemicals, S1191) were dissolved in DMSO (Sigma-Aldrich, D2438). For in vivo studies, palbociclib (MedChemExpress, HY-50767), ribociclib (MedChemExpress, HY-15777), and abemaciclib (MedChemExpress, HY-16297A) were prepared in corn oil (Spectrum Chemical, CO136) containing 10% DMSO. 5-Ethynyl-2′-deoxyuridine (EdU; Sigma-Aldrich, 900584) and AFDye-647 picolyl azide (Click Chemistry Tools, 1300) were used to label DNA synthesis.

Antibodies

Primary antibodies were from Cell Signaling Technology: Rb (9309), phospho-Rb (Ser807/811; 8516), and c-Myc (5605); from Abcam: CDK6 (ab124821), cyclin E (ab32103), and cyclin A (ab32386); and from BD Biosciences: Rb (554136). The BD Biosciences Rb antibody was used for Rb visualization in cells expressing the CDK4/6 sensor. Secondary antibodies for immunofluorescence were Alexa Fluor 488 goat anti-mouse (Thermo Fisher Scientific, A32723) and Alexa Fluor 568 goat anti-rabbit (Thermo Fisher Scientific, A11036). For immunoblotting, IRDye secondaries were used: goat anti-mouse 800CW (LI-COR, 926-32210) and goat anti-rabbit 680RD (LI-COR, 926-68071).

DNA constructs and cell line generation

DNA constructs were generated as described previously (8,35,42,43,74). Briefly, Gibson assembly was used to clone H2B-iRFP670-p2a-mCerulean-Cdt1 (1–100) (Addgene, 223965), H2B-iRFP670-p2a-mCerulean-Geminin (1–110) (Addgene, 223959), and DHB (995–1087)-mVenus-p2a-mCherry-Rb (886–928) (Addgene, 126679) into pLenti-IRES vectors bearing puromycin, blasticidin, or neomycin selection markers. Doxycycline-inducible c-Myc, cyclin E1, and cyclin A2 constructs were generated by inserting PCR products into pCW57.1 (Addgene, 50661) after NheI/BamHI digestion.

Stable cell lines were created by lentiviral transduction. Lentiviral plasmids and packaging vectors pMDLg/pRRE (Addgene, 12251), pRSV-Rev (Addgene, 12253), and pCMV-VSV-G (Addgene, 8454) were transfected into HEK-293T cells using polyethyleneimine in Opti-MEM (Thermo Fisher Scientific, 31985070). Viral supernatants were collected at 72 and 96 h, pooled, clarified (1,200 rpm, 5 min), filtered (0.45 µm; Millipore, SLHA033SB), concentrated (Amicon Ultra-15; Millipore, UFC910024; 4,000 rpm, 10 min), and stored at −80 °C. Target cells were infected in the presence of polybrene (5 µM) for 48 h and selected with puromycin (1 µg/mL; InvivoGen, ant- pr), blasticidin (10 µg/mL; InvivoGen, ant-bl), or neomycin (800 µg/mL; Thermo Fisher Scientific, BP673-5), or by FACS based on introduced fluorescent reporters.

CDK6 and Rb knockout cell lines

Rb knockout in MCF-7 cells was previously described and validated (42). For CDK6 knockout, MCF-7 cells were transfected with crRNA:tracrRNA-ATTO550 duplexes complexed with recombinant Cas9 to form RNPs. ATTO550-positive cells were single-cell sorted by flow cytometry, expanded, and screened by immunoblotting. Two gRNAs were used to target CDK6:

CDK6 gRNA 1: 5’-GACCACGUUGGGGUGCUCGAGUUUUAGAGCUAUGCU-3’

CDK6 gRNA 2: 5’-CUGGACUGGAGCAAGACUUCGUUUUAGAGCUAUGCU-3’

Drug-resistant cell lines

To generate CDK4/6i resistance, breast cancer cell lines were continuously exposed to palbociclib (1 µM) for over a month, with media and drug refreshed every 2–3 days. To establish double resistance to palbociclib and fulvestrant, palbociclib-resistant cells were then treated with fulvestrant (500 nM) for >2 months under the same refresh schedule. Resistance was confirmed by drug titration and determination of the half-maximal inhibitory concentration, calculated from the fraction of S-phase cells relative to drug-naïve controls.

Live-cell reporhters

H2B-iRFP670 was used to segment nuclei and track single cells. A fluorescent Cdt1 degron (aa 1–100) (37) reported cell-cycle phase transitions; fluorescent geminin was used to mark S-phase entry. Kinase translocation reporters (KTRs) for CDK4/6 and CDK2 (35,36) monitored kinase-dependent phosphorylation of specific substrates (Rb C-terminus [aa 886–928] for CDK4/6 and DNA helicase B [DHB; aa 994–1087] for CDK2), driving regulated nucleo-cytoplasmic shuttling. The cytoplasmic/nuclear fluorescence intensity ratio provided a quantitative readout of kinase activity. Because the CDK4/6 KTR contains a degenerate CDK2 motif and partially reports CDK2 activity in S/G2, we applied a linear-regression–derived correction based on the CDK2 reporter (35,42,44,74,75):

MCF-7: CDK4/6 activity = (CDK4/6 reporter) − 0.41 × (CDK2 reporter)

MDA-MB-231: CDK4/6 activity = (CDK4/6 reporter) − 0.35 × (CDK2 reporter)

Immunofluorescence and mRNA fluorescence in situ hybridization (FISH)

Cells were seeded in glass-bottom 96 well plates (Cellvis, P96-1.5H-N) at least 16 hr before experiments. S-phase fraction was assessed by EdU incorporation. Cells were pulsed with EdU (10 µM) for 15 min at 37°C, fixed in 4% formaldehyde (Thermo Scientific, 28906) in PBS supplemented with 10 mM HEPES (Sigma-Aldrich, H3537) for 15 min at room temperature (RT), and permeabilized in 0.2% Triton X-100 in PBS (PBS-T; Sigma-Aldrich, T8787) for 15 min. EdU was detected by click chemistry (15 min at RT) in 2 mM CuSO4 (Sigma-Aldrich, C1297), 20 mg/ml sodium ascorbate (Sigma-Aldrich, A4034), and 3 µM Alexa Fluor Dye 647 picolyl azide (Click Chemistry Tool, 1300) prepared in TBS (pH 8.3; Sigma-Aldrich, T6066).

For immunofluorescence (IF), cells were blocked (1 hr, RT) in PBS containing 10% FBS, 1% BSA, 0.1% Triton X-100, and 0.01% NaN3. Primary antibodies were incubated overnight at 4 °C in blocking buffer: anti-phospho-Rb (Ser807/811; Cell Signaling Technology, 8516), anti-Rb (BD Biosciences, 554136), and anti-c-Myc (Cell Signaling Technology, 5605). The next day, Alexa Fluor 488- or 568-conjugated secondary antibodies (Thermo Fisher Scientific) were applied for 1 hr at RT (1:2,000 in blocking buffer). Nuclei were counterstained with Hoechst 33342 (Thermo Fisher Scientific, 62249; 10 mg/mL stock diluted 1:10,000 in PBS, 15 min, RT). Cells were washed with PBS between steps and stored in PBS until imaging. For mRNA FISH, the ViewRNA ISH Cell Assay Kit (Thermo Fisher Scientific, QVC0001) and an E2F1 probe set (Thermo Fisher Scientific, VA1-12108-VC) were used according to the manufacturer’s instructions.

For experiments in Figure 3, we used a live–fixed pipeline. Asynchronously proliferating cells were imaged live for ≥48 h (37 °C, 5% CO₂; acquisition parameters detailed in “Live and fixed cell image acquisition”). Histone H2B fluorescence was used to segment/track nuclei and define time since mitosis (t = 0 at anaphase), and a CDK2 KTR provided CDK2 activity in the last live frame. Immediately after the live acquisition, the same wells were pulsed with EdU (10 µM, 15 min) and fixed/permeabilized as above. To prevent interference from fluorescent proteins during fixed assays, we photobleached residual fluorescence with 3% H₂O₂ + 20 mM HCl in PBS for 2 hr at RT. We then performed click-chemistry EdU detection, IF for phospho-Rb (Ser807/811) and total Rb, and RNA FISH for E2F1.

Growth curve measurement

To assess population growth, 5,000 cells were seeded per well in 24-well plates (Thermo Fisher Scientific, FB012929). At the indicated time points, cells were harvested by trypsinization, resuspended in 1 mL growth medium, and counted manually with a hemocytometer (Hausser Scientific, HS-3510). Cells were replated as needed for subsequent time points.

Live-cell, fixed-cell, and tumor tissue image acquisition

Cells were maintained at 30–80% confluency in 96-well glass-bottom plates (Cellvis, P96-1.5P) (75). Phenol red-free media were used to minimize background fluorescence. Imaging was performed on an inverted Nikon Eclipse Ti-2 microscope equipped with a Hamamatsu ORCA-Fusion camera. Objectives were 10× (Nikon CFI Plan Apo Lambda, NA 0.45; no binning) or 20× (Nikon CFI Plan Apo Lambda, NA 0.75; 2×2 binning). Excitation/emission settings were: CFP, 440/482 nm; GFP, 488/525 nm; YFP, 514/535 nm; mCherry, 594/609 nm; Cy5, 640/690 nm. Images were acquired using NIS-Elements AR v5.21.03. Live-cell imaging was performed in a humidified 37 °C/5% CO₂ chamber, acquiring 3 sites per well every 12 min. For fixed-cell imaging, 9 (10×) or 32 (20×) sites per well were collected. Total exposure time was kept <500 ms per time point. mRNA FISH images were collected with 1.5 µm z-stacks. Whole-tumor cross-sections were imaged on a Zeiss Axio Observer 7 with Apotome 2 and a Hamamatsu ORCA-Flash 4 camera using a 10× Zeiss Plan-Apochromat objective (NA 0.45; no binning) at 1.5 µm z-stack intervals.

Image processing and analysis

Live- and fixed-cell images were processed using custom MATLAB scripts (MathWorks, R2021a). Flat-field correction was applied to reduce illumination bias. For fixed-cell analysis, nuclei were segmented on Hoechst using a histogram-curvature threshold. For live-cell analysis, nuclei were segmented using H2B-iRFP670 with a Laplacian-of-Gaussian blob detector; adjacent nuclei were separated by marker-based watershed. Background was corrected by subtracting the 50th percentile of non-nuclear pixels. For live-to-fixed alignment, cells were segmented in the final live frame and matched to the corresponding fixed-cell field. Cell tracks were generated with a deflection-bridging algorithm. Mitoses were identified by the appearance of two daughter nuclei adjacent to a prior nucleus and by H2B intensity (each daughter about 45–55% of the mother). Cytoplasmic signal was quantified as the median intensity within a ring 0.65–3.25 µm outside the nuclear boundary; cells with overlapping rings were excluded. For RNA FISH, whole-cell areas were estimated by dilating nuclear masks by 50 µm; cells overlapping neighbors were excluded. Raw FISH images were processed with a top-hat filter (4 µm circular kernel) to generate puncta masks. Tissue images were analyzed in ImageJ.

Immunoblot

Cells were rinsed with ice-cold PBS and lysed in 300 µL CHAPS buffer (50 mM Tris-HCl, 150 mM NaCl, 1 mM EDTA, 10 mM N-ethylmaleimide, 0.3% CHAPS, 1 mM PMSF) supplemented with Halt Protease Inhibitor (Thermo Fisher Scientific, 1861279) and phosphatase inhibitor (Roche, 4906845001) for 5 min on ice. Lysates were cleared (14,000 × g, 10 min, 4 °C). Protein concentration was determined with the Pierce 660 nm assay (Thermo Fisher Scientific, 22660). Samples (12 µg) were denatured in LDS buffer (Thermo Fisher Scientific, NP0007) at 70 °C for 10 min, resolved on NuPAGE 4–12% Bis-Tris gels (Thermo Fisher Scientific, NP0322BOX), and transferred to 0.45 µm PVDF membranes (Bio-Rad, 1620174) using a Trans-Blot Turbo (Bio-Rad, 1704150) with 1× transfer buffer (Bio-Rad, 10026938). Membranes were blocked (LI-COR Intercept, 927-60001, 1 hr, RT) and incubated overnight (4 °C) with primary antibodies—Rb (CST, 9309), phospho-Rb Ser807/811 (CST, 8516), CDK6 (Abcam, ab124821), or β-actin (CST, 3700S)—diluted 1:1,000 in blocking buffer. After TBS-T washes (20 mM Tris, 150 mM NaCl, 0.1% Tween-20, pH 7.5), membranes were incubated with IRDye 800CW goat anti-mouse (LI-COR, 926-32210) or IRDye 680RD goat anti-rabbit (LI-COR, 926-68071) secondaries (1:2,000; 2 hr, RT). Blots were imaged on a LI-COR Odyssey and processed in Image Studio Lite v5.2.

RNA sequencing

Total RNA was extracted using the GeneJET RNA Purification Kit (Thermo Fisher Scientific, K0732) according to the manufacturer’s protocol. After trypsinization and PBS wash, cells were lysed in 600 µL lysis buffer supplemented with β-mercaptoethanol (Bio-Rad, 1610710), homogenized, mixed with 360 µL 100% ethanol (Fisher Scientific, BP28184), and purified on spin columns. RNA was eluted in nuclease-free water (Thermo Fisher Scientific, 10977-015). Azenta Life Sciences performed library preparation and sequencing. Count-matrix analysis was conducted in R using DESeq2 (v1.44.0) (76) with log₂ fold-change shrinkage via apeglm (77). Gene sets (Hallmark, C2, C5) were obtained from MSigDB via msigdbr (v7.5.1). Heatmaps were generated with ComplexHeatmap (78). GSEA was performed with clusterProfiler (v4.12.6) using Hallmark annotations, a maximum gene-set size of 500, and 10,000 permutations; gene sets with nominal P < 0.05 and FDR < 0.25 were considered significant.

Animal

All procedures were approved by the Institutional Animal Care and Use Committee at Columbia University Irving Medical Center and conducted in accordance with NIH guidelines. Female Foxn1-/- J:NU mice (6 weeks old; The Jackson Laboratory, 007850) were housed in a specific-pathogen-free barrier facility on a 12-hr light/dark cycle with ad libitum food and water and were monitored per institutional humane-endpoint policies.

Xenograft experiments

Female J:NU mice were anesthetized by intraperitoneal injection of ketamine (45 mg/kg) plus xylazine (5 mg/kg). Drug-naïve MCF-7 cells (8 × 10⁶ per mouse) were resuspended 1:1 in PBS:Geltrex LDEV-Free Reduced Growth Factor Basement Membrane Matrix (Thermo Fisher Scientific, A1413201) and injected orthotopically into the abdominal mammary fat pad. Tumors were measured twice weekly with digital calipers; volumes were calculated as (width² × length) × 0.5. When tumors reached approximately 100 mm³, mice received daily palbociclib (50 mg/kg, oral gavage). Upon regrowth to 145–155 mm³, mice were randomized (n = 5/group) to: (1) treatment discontinuation; (2) palbociclib 50 mg/kg daily; (3) abemaciclib 50 mg/kg daily; or (4) ribociclib 200 mg/kg daily for 32 days. At study end, mice were anesthetized and perfused with 1% formaldehyde prior to tumor collection.

Immunohistochemistry

After vascular perfusion with 1% formaldehyde in PBS, tumors were fixed in 1% formaldehyde (1 hr, 4 °C), washed in PBS-T, and cryoprotected in 30% sucrose (overnight, 4 °C). Samples were embedded in OCT (Thermo Fisher Scientific, 23-730-571), frozen at −80 °C, and sectioned at 40 µm onto Superfrost Plus slides (Thermo Fisher Scientific, 12-550-15). Sections were rehydrated in PBS-T, blocked in 5% normal goat serum (Jackson ImmunoResearch, 005-000-121) for 1 hr (RT), and incubated with recombinant anti-Rb (Abcam, ab181616; 1:1,000 in 5% serum/PBS-T) overnight (RT). After PBS-T washes, sections were incubated with Alexa Fluor 488 goat anti-rabbit (Thermo Fisher Scientific, A32731; 1:500 in PBS-T) for 4 hr (RT), counterstained with DAPI (Sigma-Aldrich, D9542; 1:500, 10 min), and mounted in Fluoromount-G (Invitrogen, 00-4958-02) for imaging.

Statistics and Reproducibility

Statistical analyses were performed in GraphPad Prism v10.2.0. For parametric data, unpaired two-tailed Student’s t-tests or one-way ANOVA with Tukey’s post-hoc analysis were used as appropriate; paired two-tailed t-tests were used for paired comparisons. The specific tests and sample sizes are reported in the figure legends; a comprehensive summary is provided in Supplementary Table 1. All experiments were reproduced in at least two independent biological replicates.